---
title: "fbat plot data analysis 2025"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries    
```{r}
# if !installed.packages("pak") {
#   install.packages("pak")
# }
# pak::pak("https://github.com/paudelsushil/Rfuels")

# Local package is installed from this repo (forestPluse)
install.packages("forestPluse_2.1.0.tar.gz", repos = NULL, type = "source")
library(forestPluse)


libs <- c("terra", "dplyr", "ggplot2", "readr", "tidyr", "lubridate", "stringr", "sf", "readxl", "here", "Rfuels", "forestPluse")

lapply(libs, library, character.only = TRUE)




```


# setup working direcotories and data sources
```{r}
# Set base directory
base_dir <- r"(D:\Clark University\Dominik Kulakowski - Sushil\Datasets\data_Rfabt_repo)"

# Set up analyzed and raw data sources
analyzed_data_src <- file.path("D:\\Clark University\\Dominik Kulakowski - Sushil\\Repositories\\ForDis\\analyzed_data")
# Set up raw data source
raw_data_src <- file.path("D:\\Clark University\\Dominik Kulakowski - Sushil\\Repositories\\ForDis\\raw_data")



# Set up climate data directories
tmmxM_dir <- file.path(analyzed_data_src, "climate_data", "tmmx", "monthly")
tmmnM_dir <- file.path(analyzed_data_src, "climate_data", "tmmn", "monthly")
prM_dir <- file.path(analyzed_data_src, "climate_data", "pr", "monthly")

pdsiM_dir <- file.path(analyzed_data_src, "climate_data", "pdsi")
scPdsiM_dir <- file.path(analyzed_data_src, "climate_data", "scpdsi")
pzIM_dir <- file.path(analyzed_data_src, "climate_data", "pzi")

vpdM_dir <- file.path(analyzed_data_src, "climate_data", "vpd", "monthly")
petM_dir <- file.path(analyzed_data_src, "climate_data", "pet", "monthly")
vsM_dir <- file.path(analyzed_data_src, "climate_data", "vs", "monthly")
rmaxM_dir <- file.path(analyzed_data_src, "climate_data", "rmax", "monthly")
rminM_dir <- file.path(analyzed_data_src, "climate_data", "rmin", "monthly")
sradM_dir <- file.path(analyzed_data_src, "climate_data", "srad", "monthly")

# Set up topographic data directory
topo_data_dir <- file.path(analyzed_data_src, "topographic_data")

```


# 1. Complining and compairing coordinates from different gpx waypoints files
```{r}
raw_gps_dir <- file.path(base_dir, "GPS_data_backup", "Garmin GPSMAP 65", "Garmin", "GPX")

gpx_files <- list.files(raw_gps_dir, pattern = "^Waypoints_.*[JUN, JUL].*-25.*\\.gpx$", full.names = TRUE)
gpx_files




gps_coords <- lapply(gpx_files, function(file) {
      gpx_data <- sf::st_read(file, layer = "waypoints", quiet = TRUE)
      coords <- sf::st_coordinates(gpx_data)
      data.frame(
        file = basename(file),
        name = gpx_data$name,
        lon = coords[, "X"],
        lat = coords[, "Y"],
        elev = gpx_data$ele,
        date = gpx_data$time
      )
})
gps_coords_df <- do.call(rbind, gps_coords)
# View(gps_coords_df)

writexl::write_xlsx(gps_coords_df, file.path(base_dir, "data", "GPS_data_backup", "compiled_gpx_data_2025", "gpx_2025_raw.xlsx"))

plots_measured_2025_sheet <- readxl::read_excel(file.path(base_dir, "data", "plots with GPS Location.xlsx"))

st_write(gps_coords_df, file.path(base_dir, "data", "GPS_data_backup", "compiled_gpx_data_2025", "gpx_2025_raw.kml"), driver = "kml", delete_dsn = TRUE)

st_write(gps_coords_df, file.path(base_dir, "data", "GPS_data_backup", "compiled_gpx_data_2025", "gpx_2025_raw.gpkg"), driver = "GPKG", delete_dsn = TRUE)

```

## Finalize GPS waypoints measured in 2025

```{r}
# View the structure of plots_measured_2025_sheet
plot_in_sheets <- plots_measured_2025_sheet |> 
  select(`Plot id`, `GPS WPT`, Latitude, Longitude) |> 
  rename(gps_wpt = `GPS WPT`, plot_id = `Plot id`)

View(plot_in_sheets)


# Extract unique GPS waypoints from fbat_fuel_depths
fbat_gps_wpts_measured <- fbat_fuel_depths |> 
  select(FBAT_plot_id, gps_wpt ) |> 
  distinct()

View(fbat_gps_wpts_measured)
# Match with GPS coordinates
# Convert gps_wpt to character to match name column in gps_coords_df
fbat_gps_wpts_final <- fbat_gps_wpts_measured |> 
  mutate(gps_wpt = as.character(gps_wpt)) |> 
  left_join(
    gps_coords_df |> 
      select(name, lon, lat, elev, date, file) |> 
      distinct(),
    by = c("gps_wpt" = "name"),
    relationship = "many-to-one"
  ) |> 
  left_join(
    plots_measured_2025_sheet,
    by = c("FBAT_plot_id" = names(plots_measured_2025_sheet)[1]),  # Match by first column (likely plot ID)
    relationship = "many-to-one"
  ) |> 
  rename(
    wpt_name = gps_wpt,
    wpt_lon = lon,
    wpt_lat = lat,
    wpt_elev = elev,
    wpt_date = date,
    gps_source = file
  )

glimpse(fbat_gps_wpts_final)
View(fbat_gps_wpts_final)

# Save finalized GPS waypoints
writexl::write_xlsx(fbat_gps_wpts_final, 
                     file.path(base_dir, "data", "analyzed_data", "fbat_gps_waypoints_2025_final.xlsx"))

```


# Make a track of navigation through all measured FBAT plots in 2025
```{r}
track_dir <- file.path(base_dir, "GPS_data_backup", "Garmin GPSMAP 65", "Garmin", "GPX", "Archive")
track_files <- list.files(track_dir, pattern = "^2025.*\\Auto.gpx$", full.names = TRUE)
track_files
tracks <- lapply(track_files, function(file) {
              gpx_data <- sf::st_read(file, layer = "tracks", quiet = TRUE)
                  }
)
track_data <- do.call(rbind, tracks)

sf::st_write(track_data, file.path(base_dir, "data", "GPS_data_backup", "compiled_gpx_data_2025", "fbat_2025_navigation_track.gpkg"), driver = "GPKG", delete_layer = TRUE)

sf::st_write(track_data, file.path(base_dir, "data", "GPS_data_backup", "compiled_gpx_data_2025", "fbat_2025_navigation_track.kml"), driver = "kml", delete_layer = TRUE)


```


# Final FBAT plot GPS coordinates for 2025
```{r}
fbat_plots <- read_excel(file.path(base_dir, "data", "analyzed_data", "fbat_plots_original.xls"))
names(fbat_plots             )

# make it a spatial data frame
fbat_plots_2025 <- read.csv(file.path(base_dir, "data", "analyzed_data", "final_plots_measured_2025.csv")) |> 
                 left_join(
                  fbat_plots, 
                  by = c("fbat_id_2025" = "fbat_plot_id")
                 )

View(fbat_plots_sf)


```




# Extract orthophotos from drone data zip files
```{r}
zip_src <- "I:/Drone_data_analysis_2025"
outpur_dir <- "I:/Drone_data_analysis_2025/drone_Orthophotos"
file_src <- "odm_orthophoto/odm_orthophoto.tif"


extract_and_rename_tifs(
  zip_dir = zip_src,
  output_dir = outpur_dir,
  file_to_extract = file_src
)


```

# Extract ndvi tif from drone data zip files
```{r}

```



## Load functions

```{r }
# Functions are provided by the locally installed forestPluse package.
packageVersion("forestPluse")

```

## Load data


```{r pressure, echo=FALSE}

fbat_crown_cover <- read_csv(file.path(base_dir, "data", "analyzed_data", "fbat_crown_cover.csv"))
fbat_tree_measurement <- read_csv(file.path(base_dir, "data", "analyzed_data", "fbat_tree_measurement.csv"))
fbat_vegetation <- read_csv(file.path(base_dir, "data", "analyzed_data", "fbat_vegetation.csv"))
fbat_fuel_depths <- read_csv(file.path(base_dir, "data", "analyzed_data", "fbat_fuel_depths.csv"))

glimpse(fbat_crown_cover)
glimpse(fbat_tree_measurement)
glimpse(fbat_vegetation)
View(fbat_fuel_depths)


```

## Match GPS waypoints from fuel depths with GPS coordinates

```{r}
# Match gps_wpt from fbat_fuel_depths with name column in gps_coords_df
fbat_fuel_depths_with_coords <- fbat_fuel_depths |> 
  left_join(
    gps_coords_df |> 
      select(name, lon, lat, elev, date) |> 
      distinct(),
    by = c("gps_wpt" = "name"),
    relationship = "many-to-one"
  ) |> 
  rename(
    gps_wpt_lon = lon,
    gps_wpt_lat = lat,
    gps_wpt_elev = elev,
    gps_wpt_date = date
  )

glimpse(fbat_fuel_depths_with_coords)
View(fbat_fuel_depths_with_coords)


```

## Data exploration
```{r}
# species code reference
species_codes <- read_excel("data\\v9-5_2024-10_Natl_MasterTreeSpeciesList.xlsx")

View(species_codes)

```

# Compiling coordiantes of fbat plots measured in 2025

```{r}
fbat_measured_ipad_2025 <- data.frame(
  FBAT_plot_id = unique(fbat_fuel_depths$FBAT_plot_id),


  latitude = c(38.12345, 38.23456, 38.34567),
  longitude = c(-120.12345, -120.23456, -120.34567)
)
```

## Calculations suggested by Brown 1974
```{r}
# Slope correction function
slope_correction <- function(slope){
    angle <- slope # angle in degrees
    percent_slope <- tan(angle * (pi / 180)) * 100 # convert angle to percent slope
    corrected_slope <- sqrt(1 + (percent_slope / 100)^2) # slope correction factor
  return(round(corrected_slope, 2))
}

# QMD (Quadratic Mean Diameter) function
# "This is the diameter of the tree of average basal area. FOr a stand with only one tree, QMD = DBH.
# For a stand with more than one tree, QMD is calculated as the square root of the mean of the squared 
# diameters of all trees in the stand."
qmd <- function(dbh){ 
  qmd_value <- sqrt(mean(dbh^2, na.rm = TRUE)) # calculate QMD
  return(round(qmd_value, 2))
}

# SDI (Stand Density Index) function
sdi <- function(ba, qmd){
  sdi_value <- (ba / (0.005454 * qmd^2)) # calculate SDI
  return(round(sdi_value, 2))
} 


slope_correction(30) # example usage                    
```     

## Calculate plot wise canopy density
```{r}

crwonDensity <- calculate_crownDensity(df = fbat_crown_cover)


```

## Calculate plot wise basal area, QMD, SDI
```{r}
# Basal area per plot and species
fbat_trees <- fbat_tree_measurement |> 
  select(FBAT_plot_id, tree_tag, tree_species, tree_dbh, tree_height)
fbat_spps_basalArea <- calculate_plot_species_basal_area(
                                          ,
                                          dbh_col = "tree_dbh",
                                          species_col = "tree_species",
                                          plot_col = "FBAT_plot_id",
                                          height_col = "tree_height",
                                          units = "m2"
                                          )
fbat_spps_basalArea

# Calculate QMD and SDI per plot
fbat_plot_tree <- fbat_spps_basalArea |> 
  mutate(
    QMD = qmd(dbh = mean_dbh_cm),
    SDI = sdi(ba = total_basal_area_m2, qmd = QMD)
  )

View(fbat_plot_tree)
```

# 1. Post-disturbance regeneration and fuel depths
 ## Do topography and landscape-level patterns in fire severity predict post-disturbace regeneration 
 ## following high-severity fire ? 


```{r}
ca_GISdata_dir <- "D:\\Datasets"


# Load fuel management data
fuel_mgmt <- st_read(file.path(ca_GISdata_dir, "CA_fuelMgmt.gpkg")) |> 
                st_transform(st_crs("epsg:5070"))


# Load wildfire data
fire <- st_read(file.path(ca_GISdata_dir, "CA_mtbs_perims.gpkg")) |> 
                st_transform(st_crs("epsg:5070"))
# Load outbreak data
outbreaks <- st_read(file.path(ca_outbreak_src, "cal_outbreaks.gpkg"))  |> 
                st_transform(st_crs("epsg:5070"))

```

# FBAT plots
```{r}
# Total plots measured in 2025
# fbat_plots_2025 <- read_excel((here::here("Rfbat/data/fbat_measured in 2025 with details.xls")), sheet = "final_plots_2025")
fbat_plots_2025 <- sf::st_read(here::here("Rfbat\\data\\analyzed_data\\fbat_plots_2025.gpkg")) |> 
                    st_transform(st_crs("epsg:5070")) |> 
                    select(-c(Join_Count, TARGET_FID, Field, t1_y, t1_x, t1_azmth, t1_slope)) |> 
                    mutate(
                      x = st_coordinates(Shape)[,1],
                      y = st_coordinates(Shape)[,2]
                    )
View(fbat_plots_2025)
```


# Preparing Regeneration data for analysis
```{r}
# Vegetation data from iPads Pro
regeneration_ipadsPro <- read.csv(here::here("Rfbat/data/iPad Pro data/vegetation_pro.csv")) |> 
                        select(
                        FBAT_plot_id, transect, transect_slope, transect_azimuth, 
                        herb_pcnt_alive_est,	herbs_avg_hgt_est,	herbs_pcnt_cover_est,	
                        grass_avg_hgt_est,	grass_pcnt_alive_est,	grass_pcnt_cover_est,
                        shrubs_avg_hgt_est,	shrubs_pcnt_alive_est,	shrubs_pcnt_cover_est,
                        seedlingsSaplings_avg_hgt_est,	seedlingsSaplings_pcnt_alive_est,	seedlingsSaplings_pcnt_cover_est,	seedling_sapling_count,
                        species, seedling_sapling_type, height_cm, dbh, alive_dead, vigore_class
                        ) |> 
                        mutate(
                          transect = case_when(
                          transect == "Transect 1" ~ 1,
                          transect == "Transect 2" ~ 2,
                          transect == "Transect 3" ~ 3)
                          ) |>
                        filter(
                          !is.na(FBAT_plot_id)
                          )

View(regeneration_ipadsPro)

# Vegetation data from Sushil's iPad
regeneration_Myipad <- read.csv(here::here("Rfbat/data/Sushil's iPad/vegetation_sp.csv")) |> 
                        select(
                        FBAT_plot_id, transect, transect_slope, transect_azimuth,
                        herb_pcnt_alive_est,	herbs_avg_hgt_est,	herbs_pcnt_cover_est,	
                        grass_avg_hgt_est,	grass_pcnt_alive_est,	grass_pcnt_cover_est,
                        shrubs_avg_hgt_est,	shrubs_pcnt_alive_est,	shrubs_pcnt_cover_est,
                        seedlingsSaplings_avg_hgt_est,	seedlingsSaplings_pcnt_alive_est,	seedlingsSaplings_pcnt_cover_est,	seedling_sapling_count,
                        species, seedling_sapling_type, height_cm, dbh, alive_dead, vigore_class
                        ) |> 
                        mutate(
                          transect = case_when(
                          transect == "Transect 1" ~ 1,
                          transect == "Transect 2" ~ 2,
                          transect == "Transect 3" ~ 3)
                          ) |>
                        filter(
                          !is.na(FBAT_plot_id)
                          )
View(regeneration_Myipad)


# Vegetation data from scanned datasheets
regeneration_datasheet <- read_excel(
                                  (here::here("Rfbat/data/FBAT_plot_data_2025.xlsx")),
                                  sheet = "vegetation_regeneration"
                                  ) |> 
                                  select(
                                  FBAT_plot_id, transect, transect_slope, transect_azimuth,
                                  herb_pcnt_alive_est,	herbs_avg_hgt_est,	herbs_pcnt_cover_est,	
                                  grass_avg_hgt_est,	grass_pcnt_alive_est,	grass_pcnt_cover_est,
                                  shrubs_avg_hgt_est,	shrubs_pcnt_alive_est,	shrubs_pcnt_cover_est,
                                  seedlingsSaplings_avg_hgt_est,	seedlingsSaplings_pcnt_alive_est,	seedlingsSaplings_pcnt_cover_est,	seedling_sapling_count,
                                  species, seedling_sapling_type, height_cm, dbh, alive_dead, vigore_class
                                  ) |> 
                                  filter(
                                    !is.na(FBAT_plot_id)
                                  ) |> 
                                  mutate(
                                    seedlingsSaplings_pcnt_alive_est = as.numeric(
                                      ifelse(is.na(seedlingsSaplings_pcnt_alive_est), 0.0, seedlingsSaplings_pcnt_alive_est)
                                      ),
                                    seedlingsSaplings_pcnt_cover_est = as.numeric(
                                      ifelse(is.na(seedlingsSaplings_pcnt_cover_est), 0.0, seedlingsSaplings_pcnt_cover_est)
                                      )
                                  )
                                  

View(regeneration_datasheet)

# Verify total columns and column names in each dataset
View(janitor::compare_df_cols(
                          regeneration_ipadsPro,
                          regeneration_Myipad,
                          regeneration_datasheet
                          )
)


# Combine all regeneration datasets into one
regeneration_combined <- bind_rows(
                          regeneration_ipadsPro,
                          regeneration_Myipad,
                          regeneration_datasheet
                          )


```



## Prepare tree_measurement data for analysis
```{r}
# Tree measurement data from iPads Pro
tree_measurement_ipadsPro <- read.csv(here::here("Rfbat/data/iPad Pro data/tree_measurement_pro.csv")) |> 
                            select(FBAT_plot_id, tree_tag, tree_azm, tree_distance, tree_species, tree_dbh, tree_height, tree_cbh, tree_hlc, deadAlive, tree_vigore_class, mortality_phase,	snag_class,	pitch_tubes,	frass,	bb_gallaries,	bb_spps,	fire_damage,	mechanical_damage,	suppressed,	tree_disease,	cod,	THM_comment,	tree_comment) |>
                            filter(!is.na(FBAT_plot_id))
View(tree_measurement_ipadsPro) 

# Tree measurement data from Sushil's iPad
tree_measurement_Myipad <- read.csv(here::here("Rfbat/data/Sushil's iPad/tree_measurement_sp.csv")) |> 
                            select(FBAT_plot_id, tree_tag, tree_azm, tree_distance, tree_species, tree_dbh, tree_height, tree_cbh, tree_hlc, deadAlive, tree_vigore_class, mortality_phase, snag_class,	pitch_tubes,	frass,	bb_gallaries,	bb_spps,	fire_damage,	mechanical_damage,	suppressed,	tree_disease,	cod,	THM_comment,	tree_comment) |>
                            filter(!is.na(FBAT_plot_id))
View(tree_measurement_Myipad)

# Tree measurement data from scanned datasheets
tree_measurement_datasheet <- read_excel(
                                  (here::here("Rfbat/data/FBAT_plot_data_2025.xlsx")),
                                  sheet = "tree_measurement"
                                  ) |>
                                  select(FBAT_plot_id, tree_tag, tree_azm, tree_distance, tree_species, tree_dbh, tree_height, tree_cbh, tree_hlc, deadAlive, tree_vigore_class, mortality_phase, snag_class,	pitch_tubes,	frass,	bb_gallaries,	bb_spps,	fire_damage,	mechanical_damage,	suppressed,	tree_disease,	cod,	THM_comment,	tree_comment) |>
                                  filter(!is.na(FBAT_plot_id))
View(tree_measurement_datasheet)

# Verify total columns and column names in each dataset
View(janitor::compare_df_cols(
                          tree_measurement_ipadsPro,
                          tree_measurement_Myipad,
                          tree_measurement_datasheet
                          )
)

# Combine all tree measurement datasets into one
tree_measurement_combined <- bind_rows(
                          tree_measurement_ipadsPro,
                          tree_measurement_Myipad,
                          tree_measurement_datasheet
                          ) 



# Join tree measurement data with fbat plot data
tree_measurement_combined <- tree_measurement_combined |> 
                            left_join(
                              fbat_plots_2025 |> 
                            st_drop_geometry() |>  
                               select_all(),                                 
                              by = c("FBAT_plot_id" = "fbat_id_2025")
                            )

View(tree_measurement_combined)







```

# Locate the tree and calculate coordinate of trees based on plot GPS coordinates, tree azimuth and distance from plot center
```{r}
tree_coordinate <- locate_tree(
                    x = tree_measurement_combined$x,
                    y = tree_measurement_combined$y,
                    azimuth = tree_measurement_combined$tree_azm,
                    distance = tree_measurement_combined$tree_distance,
                    id = tree_measurement_combined$FBAT_plot_id
                    )

View(tree_coordinate)

tree_measurement_combined <- tree_measurement_combined |> 
                            bind_cols(tree_coordinate)

# save tree_measurement_combined for further analysis
writexl::write_xlsx(tree_measurement_combined, 
              file.path(base_dir, "data", "analyzed_data", "tree_measurement_final_2025.xlsx"))

# Final cleaned tree measurement data table 
tree_measurement_combined <- read_excel(file.path(base_dir, "data", "analyzed_data", "tree_measurement_final_2025.xlsx"))
View(tree_measurement_combined)



```

# comparing tree height measured in field and from crown height model
```{r}

```


# Preparing crown cover data for analysis
```{r}
# Crown cover data from iPads Pro
crown_cover_ipadsPro <- read.csv(here::here("Rfbat/data/iPad Pro data/crown_cover_pro.csv")) |> 
                        select(FBAT_plot_id, crown_cover_est
```

# Standardize species names in the combined dataset
```{r}
regeneration_combined <- clean_species_names(
                          df = regeneration_combined,
                          species_col = species
                          )
View(regeneration_combined)
colnames(regeneration_combined)
```


# 
```{r}
create_species_pivot <- function(data, 
                                 species_col = "species",
                                 keep_species = c("Ponderosa Pine", 
                                                  "Incense Cedar", 
                                                  "White Fir"),
                                 group_cols = c("FBAT_plot_id", "transect")) {
  
  
  # Step 1: Reclassify species — unspecified become "Others"
  data <- data %>%
    mutate(species_group = ifelse(!!sym(species_col) %in% keep_species,
                                  !!sym(species_col),
                                  "Others"))
  
  # Step 2: Summarize count per group
  summary_df <- data %>%
    group_by(across(all_of(group_cols)), species_group) %>%
    summarise(species_count = n(), .groups = "drop")
  
  # Step 3: Pivot wider so each species group becomes a column
  pivot_df <- summary_df %>%
    pivot_wider(names_from  = species_group,
                values_from = species_count,
                values_fill = 0)
  
  # Step 4: Join back non-species summary columns (first row per group)
  meta_cols <- setdiff(names(data), 
                       c(species_col, "species_group", "seedling_sapling_count",
                         "seedling_sapling_type", "height_cm", "dbh", 
                         "alive_dead", "vigore_class"))
  
  meta_df <- data %>%
    group_by(across(all_of(group_cols))) %>%
    summarise(across(all_of(setdiff(meta_cols, group_cols)), 
                     ~ first(na.omit(.))), 
              .groups = "drop")
  
  # Step 5: Merge metadata with pivot counts
  result <- left_join(meta_df, pivot_df, by = group_cols)
  
  return(result)
}

species_pivot_result <- create_species_pivot(
  data = regeneration_combined,
  species_col = "species",
  keep_species = c("Ponderosa Pine", "Incense Cedar", "White Fir"),
  group_cols = c("FBAT_plot_id", "transect")
)

View(species_pivot_result)

```  




# Join these information with FBAT plot information with outbreak, wildfire, and fuel management data to explore patterns in post-disturbance regeneration
```{r}
# Join species pivot result with fbat_plots_2025fb
spatialJoin_table <- readxl::read_xlsx(here::here("Rfbat\\data\\analyzed_data\\fbat_spatialJoin_fire_fuel_outbreaks.xlsx"))
View(spatialJoin_table)

fbat_plots_with_regeneration <- spatialJoin_table |> 
  left_join(species_pivot_result, by = c("fbat_id_2025" = "FBAT_plot_id"))

View(fbat_plots_with_regeneration)

# Export fbat_plots_with_regeneration for further analysis  
writexl::write_xlsx(fbat_plots_with_regeneration, 
                     file.path(base_dir, "data", "analyzed_data", "fbat_plots_with_regeneration_2025.xlsx"))


fbat_plots_with_regeneration <- readxl::read_xlsx(file.path(base_dir, "data", "analyzed_data", "fbat_plots_with_regeneration_2025.xlsx"))
View(fbat_plots_with_regeneration)
```


# Prepare final approved fbat plots list
```{r}
# Matched approved FBAT plots from regeneration plots and extract relevant GPS data
gps_backup_dir <- file.path("D:\\Datasets\\Field data 2025 backup\\GPS_data_backup")

compiled_gpx_2025 <- readxl::read_xlsx(
  file.path(gps_backup_dir, "Field_plot_progress.xlsx"),
  sheet = "compiled_gpx_2025"
) |>
  mutate(gps_wpt = readr::parse_number(as.character(name))) |>
  select(gps_wpt, date) |>
  distinct(gps_wpt, .keep_all = TRUE)

View(compiled_gpx_2025)


approved_fbat_plots <- readxl::read_xlsx(
  file.path(gps_backup_dir, "approved_field_plots_2026.xlsx"),
  sheet = "final_approved_plots"
) |>
  mutate(gps_wpt = readr::parse_number(as.character(gps_wpt)))

View(approved_fbat_plots)

# Matched the gps_wpt in approved_fbat_plots with gps_wpt in compiled_gpx_2025 and get date information from compiled_gpx_2025 into Measured_date in approved_fbat_plots
fbat_plots_approved <- approved_fbat_plots |>
  left_join(compiled_gpx_2025, by = "gps_wpt") |>
  rename(measured_fulldate = date) |> 
  mutate(Measured_date = date(measured_fulldate))

View(fbat_plots_approved)


fbat_plots_with_regeneration <-  readxl::read_xlsx(file.path(base_dir, "data", "analyzed_data", "fbat_plots_with_regeneration_2025.xlsx"))

sheet_transect_reentered <- readxl::read_xlsx(file.path(base_dir, "data", "analyzed_data", "Fbat_Reg_Slope added.xlsx"))


View(sheet_transect_reentered)

View(fbat_plots_with_regeneration)


# Extract t1_azmth, t2_azmth, and t3_azmth from fbat_plots_with_regeneration
transect_azimuth_regen <- fbat_plots_with_regeneration |>
  mutate(
    fbat_id_2025 = as.character(fbat_id_2025),
    transect_num = readr::parse_number(as.character(transect)),
    transect_azimuth = suppressWarnings(as.numeric(transect_azimuth))
  ) |>
  filter(
    !is.na(fbat_id_2025),
    transect_num %in% c(1, 2, 3),
    !is.na(transect_azimuth)
  ) |>
  mutate(transect_col = paste0("t", transect_num, "_azmth")) |>
  group_by(fbat_id_2025, transect_col) |>
  summarise(transect_azimuth = dplyr::first(transect_azimuth), .groups = "drop") |>
  tidyr::pivot_wider(names_from = transect_col, values_from = transect_azimuth)

# Second attempt: extract available azimuth from sheet_transect_reentered
transect_azimuth_reentered <- sheet_transect_reentered |>
  mutate(
    fbat_id_2025 = as.character(fbat_id_2025),
    transect_num = readr::parse_number(as.character(transect)),
    transect_azimuth = suppressWarnings(as.numeric(transect_azimuth))
  ) |>
  filter(
    !is.na(fbat_id_2025),
    transect_num %in% c(1, 2, 3),
    !is.na(transect_azimuth)
  ) |>
  mutate(transect_col = paste0("t", transect_num, "_azmth")) |>
  group_by(fbat_id_2025, transect_col) |>
  summarise(transect_azimuth = dplyr::first(transect_azimuth), .groups = "drop") |>
  tidyr::pivot_wider(names_from = transect_col, values_from = transect_azimuth)

ensure_az_cols <- function(df) {
  for (nm in c("t1_azmth", "t2_azmth", "t3_azmth")) {
    if (!nm %in% names(df)) df[[nm]] <- NA_real_
  }
  df
}

transect_azimuth_regen <- ensure_az_cols(transect_azimuth_regen)
transect_azimuth_reentered <- ensure_az_cols(transect_azimuth_reentered)

transect_azimuth_regen <- transect_azimuth_regen |>
  rename(
    t1_azmth_regen = t1_azmth,
    t2_azmth_regen = t2_azmth,
    t3_azmth_regen = t3_azmth
  )

transect_azimuth_reentered <- transect_azimuth_reentered |>
  rename(
    t1_azmth_reentered = t1_azmth,
    t2_azmth_reentered = t2_azmth,
    t3_azmth_reentered = t3_azmth
  )

fbat_plots_approved <- fbat_plots_approved |>
  select(-any_of(c(
    "t1_azmth", "t2_azmth", "t3_azmth",
    "t1_azmth_regen", "t2_azmth_regen", "t3_azmth_regen",
    "t1_azmth_reentered", "t2_azmth_reentered", "t3_azmth_reentered"
  ))) |>
  mutate(fbat_id_2025 = as.character(fbat_id_2025)) |>
  left_join(transect_azimuth_regen, by = "fbat_id_2025") |>
  left_join(transect_azimuth_reentered, by = "fbat_id_2025") |>
  mutate(
    t1_azmth = dplyr::coalesce(t1_azmth_regen, t1_azmth_reentered, 0),
    t2_azmth = dplyr::coalesce(t2_azmth_regen, t2_azmth_reentered, 120),
    t3_azmth = dplyr::coalesce(t3_azmth_regen, t3_azmth_reentered, 270)
  ) |>
  select(-matches("_regen$|_reentered$"))

missing_azmth <- fbat_plots_approved |>
  filter(!is.na(fbat_id_2025), is.na(t1_azmth), is.na(t2_azmth), is.na(t3_azmth)) |>
  distinct(fbat_id_2025)

if (nrow(missing_azmth) > 0) {
  message(
    "No transect azimuths found for fbat_id_2025: ",
    paste(missing_azmth$fbat_id_2025, collapse = ", ")
  )
}

View(fbat_plots_approved)

```

# Extract the table requested by Prof. Kulakowski
```{r}
# Export the table requested by Prof. Kulakowski
readr::write_csv(fbat_plots_approved, file.path(base_dir, "data", "analyzed_data", "fbat_plots_approved_2025.csv"))

# Some manual updates have been made in this table after initial export
field_plots <- readxl::read_xlsx(file.path(base_dir, "data", "analyzed_data", "fbat_plots_approved_2026.xlsx"))
View(field_plots)

# Create sf object from field_plots
field_plots_sf <- field_plots |>
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)


```

# Extract topographic information using field_plots
```{r}




```


# Extract Climatic variables using field_plots
```{r}

```




```{r}
# Find the date for following gps wpts
redmt_gpsWPT <- data.frame(gps_wpt = c(24,0,27,7,6,28,30,10,11,25,26,9,13,12,23,0,19,22,20,8,33,0,0,36,0),
                          fbat_id_2025 = c(158, 164, 166, 168, 169, 171, 173, 174, 177, 181, 183, 190, 195, 198, 199, 203, 205, 206, 209, 211, 213, 204, 167, 200, 202)
                          )

# extract date information for these gps wpts from compiled_gpx_2025
redmt_gpsWPT_dates <- compiled_gpx_2025 |>
  filter(gps_wpt %in% redmt_gpsWPT$gps_wpt) |> 
  mutate(surveydate = as.Date(date)) |> 
  mutate(fbat_id_2025 = redmt_gpsWPT$fbat_id_2025[match(gps_wpt, redmt_gpsWPT$gps_wpt)]) |>
  select(fbat_id_2025, gps_wpt, surveydate)

View(redmt_gpsWPT_dates)

# Export this table for further analysis
writexl::write_xlsx(redmt_gpsWPT_dates, file.path(base_dir, "data", "analyzed_data", "redmt_gpsWPT_dates.xlsx"))


# extract t1_azmth, t2_azmth, and t3_azmth for these gps wpts from regeneration_combined and reentered sheet
redmt_plots_info <- redmt_gpsWPT_dates |>
  left_join(
    fbat_plots_with_regeneration |> 
      select(fbat_id_2025, transect, transect_azimuth) |> 
      mutate(transect_col = paste0("t", transect, "_azmth")) |> 
      group_by(fbat_id_2025, transect_col) |>

      summarise(transect_azimuth = dplyr::first(transect_azimuth), .groups = "drop") |> 
      tidyr::pivot_wider(names_from = transect_col, values_from = transect_azimuth),
    by = "fbat_id_2025"
  ) |>
  left_join(
    sheet_transect_reentered |> 
      select(fbat_id_2025, transect, transect_azimuth) |> 
      mutate(transect_col = paste0("t", transect, "_azmth")) |> 
      group_by(fbat_id_2025, transect_col) |> 
      summarise(transect_azimuth = dplyr::first(transect_azimuth), .groups = "drop") |> 
      tidyr::pivot_wider(names_from = transect_col, values_from = transect_azimuth),
    by = "fbat_id_2025"
  ) |>
  left_join(
    sheet_transect_reentered |> 
      select(fbat_id_2025, transect, transect_azimuth) |> 
      mutate(transect_col = paste0("t", transect, "_azmth")) |> 
      group_by(fbat_id_2025, transect_col) |> 
      summarise(transect_azimuth = dplyr::first(transect_azimuth), .groups = "drop") |> 
      tidyr::pivot_wider(names_from = transect_col, values_from = transect_azimuth),
    by = "fbat_id_2025"
  ) |>
  # mutate(
  #   t1_azmth = dplyr::coalesce(t1_azmth.x, t1_azmth.y, 0),
  #   t2_azmth = dplyr::coalesce(t2_azmth.x, t2_azmth.y, 120),
  #   t3_azmth = dplyr::coalesce(t3_azmth.x, t3_azmth.y, 270)
  # ) |>
  select(fbat_id_2025, gps_wpt, surveydate, t1_azmth, t2_azmth, t3_azmth)



redmt_plots_info

```

# FBAT 

```{r}
measured_fbat_plots <- read_excel(file.path(base_dir, "data", "analyzed_data", "fbat_plots_approved_2026.xlsx"), sheet = "fbat_plot_approved_2026")

original_fbat_plots <- st_read(file.path(base_dir, "data", "analyzed_data", "measured_plots", "FBAT_Plots_Info.gpkg"))

# Create a column in original_fbat_plots that matches the fbat_id_2025 in measured_fbat_plots
original_fbat_plots <- original_fbat_plots |> 
  mutate(measured_status = ifelse(fbat_id_2025 %in% measured_fbat_plots$fbat_id_2025, "Measured", "Remaining")) |> 
  # use only name excluding "National Forest" for each values in FORESTNAME column to create a new column NF_name
  mutate(NF_name = as.factor(gsub(" National Forest", "", FORESTNAME, " NF"))) 



# Plot bar chart to show measured vs not measured plots per National FOrest
plot_counts <- original_fbat_plots |>
  st_drop_geometry() |>
  count(NF_name, measured_status, name = "count")

plot_counts_total <- plot_counts |>
  group_by(NF_name) |>
  summarise(count = sum(count), .groups = "drop") |>
  mutate(measured_status = "Total")


plot_counts_all <- bind_rows(plot_counts, plot_counts_total) |>
  mutate(measured_status = factor(measured_status, levels = c("Measured", "Remaining", "Total")))

plot_counts_summary <- plot_counts_all |>
  mutate(measured_status = as.character(measured_status)) |>
  group_by(NF_name) |>
  summarise(
    measured = sum(count[measured_status == "Measured"], na.rm = TRUE),
    remaining = sum(count[measured_status == "Remaining"], na.rm = TRUE),
    total = sum(count[measured_status == "Total"], na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(NF_name)

print(plot_counts_summary)
writexl::write_xlsx(plot_counts_summary, file.path(base_dir, "outputs", "fbat_plot_counts_summary.xlsx"))


ggplot(plot_counts_all, aes(x = NF_name, y = count, fill = measured_status)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = count),
            position = position_dodge(width = 0.9),
            vjust = -0.25,
            size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +

  scale_fill_manual(values = c("Measured" = "#2E8B57", "Remaining" = "#D95F02", "Total" = "#4D4D4D")) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  labs(title = "Measurement status of FBAT plots per national forest as of 2025 field season",
       x = "",
       y = "Count of FBAT plots",
       fill = "") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(angle = 0, hjust = 1, size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12, margin = margin(t = 0.5)),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.position = c(0.8, 1),
    legend.justification = c(0.5, 1),
    legend.direction = "horizontal",
    legend.background = element_rect(fill = "white", color = NA),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    plot.margin = margin(t = 14, r = 10, b = 10, l = 10)
  )

ggsave(file.path(base_dir, "outputs", "fbat_plot_measurement_status.png"), width = 10, height = 6, dpi = 300)

st_write(original_fbat_plots, file.path(base_dir, "data", "analyzed_data", "measured_plots", "FBAT_Plots_Info_updated.gpkg"), delete_dsn = TRUE)

```
