---
title: "Objective 4 Analysis "
author: "S. Paudel"
date: "2024-11-07"
output: html_document
editor_options: 
  chunk_output_type: console
---

This is a analysis of the iLAND model output for objective 4 of project proposal.
**Hypotheses** 
*1: fire and bark beetle severity will be less influenced by pre-disturbance fuel treatments and vegetation conditions.*
*2: regeneration of three species will be most inhibited in vulnerable biophysical settings and following fire or beetle disturbances.*


# 1. Load Libraries
```{r}
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}
pak::pak("https://github.com/paudelsushil/Rfuels")

devtools::install_github("paudelsushil/forestPulse")
# Local package is installed from this repo (forestPulse)
# install.packages("forestPulse_0.1.8.tar.gz", repos = NULL, type = "source")


libs <- c("terra", "dplyr", "ggplot2", "readr", "tidyr", "lubridate", "stringr", "sf", "readxl", "here", "Rfuels", "forestPulse")

lapply(libs, library, character.only = TRUE)


```


## 1.1 Set-up working directory
```{r}
dataset_dir <- file.path("D:/Clark University/Dominik Kulakowski - Sushil/Datasets")



``` 

# 2. Working with FIA data
```{r}
# FIA Database-California urls

plot_table <- read.csv("https://wifire-data.sdsc.edu/data/ndp/Data_challenge_2024/fia_data/CA_PLOT.csv")
condition_table <- read.csv("https://wifire-data.sdsc.edu/data/ndp/Data_challenge_2024/fia_data/CA_COND.csv")
tree_table <- read.csv("https://wifire-data.sdsc.edu/data/ndp/Data_challenge_2024/fia_data/CA_TREE.csv")
ref_spps_table <- read.csv("https://wifire-data.sdsc.edu/data/ndp/Data_challenge_2024/fia_data/REF_SPP.csv") 


#  OR, Using rFIA package to load FIA data
# All data acquired from FIA Datamart: https://apps.fs.usda.gov/fia/datamart/datamart.html 
library(rFIA)

CA_FIA_data <- rFIA::getFIA(dir = file.path(dataset_dir, "FIA"), 
                            states = "CA", nCores = 4, 
                            load = TRUE, common = FALSE
                            )

```

```{r, eval = requireNamespace("tidyverse", quietly = TRUE)}

```
