---
title: "Manuscript_v1.01"
author: "S. Paudel"
date: "2024-11-07"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A. Install and Load Packages

```{r}
if (!require("drat")) {
  install.packages("drat")
 }
drat::addRepo("cloudyr", "http://cloudyr.github.io/drat")

# List of libraries
libs <- c("here", "ranger", "caret","tidyverse", "tidyquant", "arrow", "terra", "sf", "zoo", "tmap", "data.table", "ncdf4", "gridExtra", "ggspatial", "ggtern", "patchwork", "magick", "rayshader", "AzureStor" )


# List packages that are not installed
installedLibs <- libs %in% 
  rownames(
    installed.packages()
    ) 
# Install packages if not installed
if(any(installedLibs == F)){
  install.packages(
    libs[!installedLibs]
  )
}
                   
# Laod packages
base::invisible(
  lapply(
    libs, library, character.only = T
  )
)        

```

# B. Cloud Connection

```{r}

# accountEndPoint <- "https://fordis.blob.core.windows.net/"
# sas = "sv=2022-11-02&ss=bfqt&srt=sco&sp=rwdlacupyx&se=2025-05-31T06:31:09Z&st=2025-01-03T23:31:09Z&spr=https,http&sig=huCiO3ZCUlVl80fKTyO%2B8SLZZZ4rNHzkq3A3zLDvRt8%3D"

# bolbEndPointKey <- AzureStor::blob_endpoint(accountEndPoint, sas = sas)

# # List containers
# AzureStor::list_blob_containers(bolbEndPointKey)
# # List storage container
# AzureStor::list_storage_containers(bolbEndPointKey)

# analyzedCont <- AzureStor::storage_container(bolbEndPointKey, "analyzed")

# # Create directory inside analyzed container
# AzureStor::create_blob_dir(analyzedCont, "mtbsfire")



# # Convert fire_lodgepole.gpkg to fire_lodgepole.parquet

# # create bolb container
# create_storage_container(bolbEndPointKey, "raw")
# rawCont <- AzureStor::storage_container(bolbEndPointKey, "analyzed")

# # Upload geoParquet file into mtbsfire directory of analyzed blob container
# AzureStor::storage_upload(analyzedCont,
#                          src = file.path(analyzed_src, 
#                                          "mtbs_fire_data", "fire_lodgepole.parquet"),
#                          dest = "mtbsfire/fire_lodgepole.parquet")




```


```{r}

```

# C. Duck DB for vector datasets

```{r}
# demPath <- file.path(raw_src,"topo_data/LF2020_Elev_220_CONUS/LF2020_Elev_220_CONUS/Tif/LC20_Elev_220.tif")

# demrast <- rast(file.path(analyzed_src, "topo_data/conus_dem_1000m.tif"))
 
# # Convert DEM to matrix (required for rayshader)
# dem_matrix <- as.matrix(demrast, wide=TRUE)

# # Create the base 3D terrain
# dem_matrix %>%
#   sphere_shade(texture = "desert") %>%
#   add_shadow(ray_shade(dem_matrix, zscale = 30), 0.5) %>%
#   add_shadow(ambient_shade(dem_matrix), 0) %>%
#   plot_3d(dem_matrix,
#           zscale = 30,
#           windowsize = c(1000, 800),
#           phi = 45,
#           theta = 45,
#           zoom = 0.75)
# glimpse(fire_lodgepole)

# # Load and process fire perimeter
# fire_rast <- rasterize(vect(fire_lodgepole), demrast)
# plot(fire_rast)
# fire_matrix <- as.matrix(fire_rast, wide=TRUE)

# # Create RGB matrix for fire overlay (red with transparency)
# fire_rgb <- array(0, dim = c(nrow(fire_matrix), ncol(fire_matrix), 4))
# fire_rgb[,,1] <- 1  # Red channel
# fire_rgb[,,4] <- ifelse(!is.na(fire_matrix), 0.5, 0)  # Alpha channel (transparency)
# # Add fire overlay
# add_overlay(fire_rgb)

# # Add finishing touches
# render_compass()
# render_scalebar(limits = c(0, 5, 10))

# # Save the render
# render_snapshot("3d_fire_map.png")


# rayshader::add_overlay()

# rgl::rglwidget()

# glimpse(fire)
```

# D. Load Working directories

```{r}
# Define Raw datasets location
raw_src <- here::here("external/raw_data")
# Define analyzed datasets location
analyzed_src <- here::here("external/analyzed_data")

# Output files source
results_src <- here::here("results") 
fig_src <- here::here("results/figs")
tbl_src <- here::here("results/tbls")
pdf_src <- here::here("results/pdfs")


tmmxM_dir <- here::here("external/analyzed_data/climate_data", "tmmx", "monthly")
tmmnM_dir <- here::here("external/analyzed_data/climate_data", "tmmn", "monthly")
prM_dir <- here::here("external/analyzed_data/climate_data", "pr", "monthly")

pdsiM_dir <- here::here("external/raw_data/climate_data", "pdsi")
scPdsiM_dir <- here::here("external/raw_data/climate_data", "scpdsi")
pzIM_dir <- here::here("external/raw_data/climate_data", "pzi")

vpdM_dir <- here("external/analyzed_data/climate_data/vpd/monthly")
petM_dir <- here("external/analyzed_data/climate_data/pet/monthly")
vsM_dir <- here("external/analyzed_data/climate_data/vs/monthly")
rmaxM_dir <- here("external/analyzed_data/climate_data/rmax/monthly")
rminM_dir <- here("external/analyzed_data/climate_data/rmin/monthly")
sradM_dir <- here("external/analyzed_data/climate_data/srad/monthly")


```

# E. Load Functions

```{r}
source(here::here("R/functions.R"))
# source(here::here("R/getGridmet.metadata.R"))
# source(here::here("R/getGridmet.data.R"))
# source(here::here("R/gridmetDaily2Monthly.R"))
source(here::here("R/calculate.timeclass.R"))
source(here::here("R/extractValue2Points.R"))

source(here::here(r"(R\theme.R)"))
source(here::here(r"(R\corrPlot.R)"))
# source(here::here(r"(R\raster2table.R)"))

```

# 1. Load datasets

We are focused in Lodgepole Pine distribution with Western Contiguous United States.

## 1.1 Fire

```{r Fire Affected Area}
# Load a fire within the study area
fireLodgepole <- st_read(file.path(analyzed_src, 
                                    "mtbs_fire_data/fireLodgepole.gpkg"))


# fire_lodgepole <- fire_lodgepole %>% 
#                   dplyr::select(event_id, state, fire_date, fire_year, geom) %>% 
#                   mutate(FireA_ha = calculate.area(., area.unit = "ha")) %>% 
#                   calculate.timeclass(., "fire_year", 1984, 2023, 10, "fire")


# fire_parquet <- fire_lodgepole %>% 
#   mutate(geometry_wkt = st_as_text(geom)) %>% 
#   st_drop_geometry(.)



# write_parquet(fire_parquet, 
#               file.path(analyzed_src, "mtbs_fire_data", "fire_lodgepole.parquet"))
             

file.exists(file.path(analyzed_src, "mtbs_fire_data", "fireLodgepole.gpkg"))            


# Calculate area in Hactare
# Read the parquet file and convert back to SF object
# fire_lodgepole_restored <- open_dataset(file.path(analyzed_src, "mtbs_fire_data/fire_lodgepole.parquet")) %>%
#   collect() %>%
#   st_as_sf(wkt = "geometry_wkt", crs = st_crs("EPSG:5070"))


fireLodgepole

```

### 1.1.1 Fire severity

```{r Fire Severity, eval = FALSE}
# Create a centriod for each fire
# fireCentroids <- lapply(1:nrow(fire_lodgepole), function(fireRow){
#   fireCentroids <- sf::st_centroid(fire_lodgepole[fireRow, ]) %>% 
#     mutate(centroid_X = st_coordinates(.)[1],
#          centroid_Y = st_coordinates(.)[2])
  
#     return(fireCentroids)
#   }) 
# # combined by row
# fireCentroids <- do.call(rbind, fireCentroids)

# # write a file
# st_write(fireCentroids, file.path(here::here("external/analyzed_data/mtbs_fire_data/mtbs_severity", "fireCentroids.gpkg")), append = F)

# # Fire Severity Info
# fireSeverityInfo <- st_read(file.path(here::here("external/analyzed_data/mtbs_fire_data/mtbs_severity", "fireCentroidsMtbsI.gpkg")))
```

### Figure8

```{r fire severity plot}
# plot_a <- plot.fireSeverity(fireSeverityInfo, periodic = F)
# plot_b <- plot.fireSeverity(fireSeverityInfo, periodic = T)


# # Combine the plots
# combined_plot <- grid.arrange(plot_a, plot_b,
#                               ncol = 1,
#                               heights = c(1, 1))
# # Save the plot
# ggsave(file.path(fig_src, "figure8.png"), combined_plot,
#        width = 7, height = 6, dpi = 300)


install.packages("fmsb")

library(fmsb)

ls('package:fmsb')
```

## 1.2 Fuel

```{r Fuel Management area}
fuel_lodgepole <- st_read(file.path(analyzed_src, 
                                    "fuel_mgmt/fuel_lodgepole.gpkg"))

# fuel_lodgepole %>% filter(is.na(JFSP_CODE) & fuel_present == 1)

# Calculate area in Hactare
fuel_lodgepole <- fuel_lodgepole %>% 
  dplyr::select(state_abbr, fuel_date, fuel_year, JFSP_CODE, activity, geom) %>% 
  mutate(FuelA_ha = calculate.area(., area.unit = "ha"),
         fuelMgmt = "Managed") %>% 
  calculate.timeclass(., "fuel_year", 1984, 2023, 10, "fuel") 
  
  
totalManagedA_km <-  sum(fuel_lodgepole$FuelA_ha)

fuel_lodgepole
totalManagedA_km
```

## 1.3 Outbreak

```{r Outbreak Area}
outbreak_lodgepole <- st_read(file.path(analyzed_src, 
                                        "outbreaks_data/outbreak_lodgepole.gpkg"))
outbreak_lodgepole <- outbreak_lodgepole %>% 
  st_transform(crs = "epsg:5070") %>% 
  mutate(OutbreakA_ha = calculate.area(., "ha")) %>% 
  calculate.timeclass(., "outbreak_year", 1984, 2023, 10, "outbreak")


outbreak_lodgepole
```

## 1.4 Study area

### 1.4.1 Administrative Boundary of 11 Western Conus

```{r Administrative boundary}
conus_states <- st_read(file.path(analyzed_src, 
                        "conus_states/conus_states.gpkg"),
                        layer = "conus_states")

# sum(conus_states$states_area_km2) # 3072131 km2
```

### 1.4.2 Lodgepole Pine (P. contorta) dominated forest area

```{r Study area}
study_area <- st_read(file.path(analyzed_src, 
                                "study_area/lodgepoleExtent.gpkg"))

study_area <- study_area %>% 
    mutate(studyA_ha = calculate.area(., "ha")) %>% 
  st_transform(., "epsg:5070")



# sum(study_area$studyA_ha) # 36765700 ha

studyA_ha <- sum(study_area$studyA_ha) # 367657 km2
studyA_ha
```

## 1.5 Outbreak between 1984 & 1996

```{r Outbreak before 1996, eval = FALSE}

ads78_04 <- st_read(file.path(raw_src, "outbreaks_data/r5/ADS_Regionwide_78_04.gdb"), layer = "ADS_78_04")

adsr6 <- st_read(file.path(analyzed_src, "outbreaks_data/r6/R6_1984_1996.gpkg"))


agent_list <- st_read(file.path(raw_src, "outbreaks_data/r5/ADS_Regionwide_78_04.gdb"), layer = "damage_agent")

agent_list[agent_list$COMMON_NAM == "mountain pine beetle", ]

sort(agent_list$COMMON_NAM)

ads78_04 <- ads78_04 %>% rename(CODE = dca_common_name ) %>% 
  mutate(CODE = as.numeric(CODE))
  
ads78_04 <- right_join(x = ads78_04, y = agent_list, by = "CODE", copy = T)



r4 <- ads78_04 %>% dplyr::select(RPT_YR, COMMON_NAM, Shape) %>% 
  filter(RPT_YR %in% c("1994","1995", "1996") & COMMON_NAM == "mountain pine beetle")

r4 <- r4 %>% rename(outbreak_year = RPT_YR, 
                    dca_common = COMMON_NAM)
adsr6 <- adsr6 %>% rename(outbreak_year = outbreak_y)

outbreak1984_1996 <- st_transform(r4, "epsg:5070") %>% 
  rbind(., st_transform(adsr6, "epsg:5070"))

outbreak1984_1996 <- outbreak1984_1996 %>% rename(outbreak_agents = dca_common)
outbreak1984_1996 <- outbreak1984_1996 %>% rename(geom = Shape)

outbreak_conus <- st_read(file.path(outbreak_src, "outbreak_conus.gpkg"))

table(outbreak1984_1996$outbreak_agents)


outbreak1984_2023 <- outbreak_conus %>% 
  dplyr::select(outbreak_year, outbreak_agents, geom) %>%
  rbind(., outbreak1984_1996)


outbreak1984_2023 <- outbreak1984_2023 %>% 
  mutate(outbreakArea_km2 = calculate.area(., "km2"))


st_write(outbreak1984_1996, file.path(analyzed_src, "outbreaks_data/outbreak1984_1996.gpkg"))


hist(out)
```

## Figure2

```{r figure2, echo = FALSE}

# Panel A: MPB outbreaks over time
plot_a <- ggplot(outbreakLodgepole, aes(x = factor(outbreakYear), y = OutbreakA_ha)) +
  geom_bar(stat = "identity", fill = "#1f77b4", width = 0.8) +
  labs(x = NULL,  # remove x label as it's shared
       y = expression("Area (ha × 10"^"3"*")"),
       title = "(a) Mountain Pine Beetle Outbreaks") +
  plot_theme() +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = function(x) x/1000,  # Convert to thousands
                       expand = c(0, 0))
plot_a  
# Panel B: Fires over time
plot_b <- ggplot(fireLodgepole, aes(x = factor(fireYear), y = FireA_ha)) +
  geom_bar(stat = "identity", fill = "#ff7f0e", width = 0.8) +
  labs(x = NULL,  # remove x label as it's shared
       y = expression("Area (ha × 10"^"3"*")"),
       title = "(b) Fire") +
  plot_theme() +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = function(x) x/1000,  # Convert to thousands
                       expand = c(0, 0))
plot_b
# Panel C: Fuel management over time
plot_c <- ggplot(fuelLodgepole, aes(x = factor(fuelYear), y = FuelA_ha)) +
  geom_bar(stat = "identity", fill = "#2ca02c", width = 0.8) +
  labs(x = "Year",
       y = expression("Area (ha × 10"^"3"*")"),
       title = "(c) Fuel Management") +
  plot_theme() +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = function(x) x/1000,  # Convert to thousands
                       expand = c(0, 0))

# Panel D: Annual mean, maximum, and Minimum PDSI values
pdsiM_df <- raster2table_annual(pdsiM_dir, "\\.nc$", "pdsiM", aoi = conusStates)

vpdM_df <- raster2table_annual(vpdM_dir, "\\.nc$", "vpdM", aoi = conusStates)


plot_d <- plot_climate_trends(pdsiM_df, "PDSI")

plot_d
# Combine the plots
combined_plot <- grid.arrange(plot_a, plot_b, plot_c, plot_d,
                            ncol = 2, 
                            heights = c(1, 1))


# Save the plot
ggsave(file.path(fig_src, "figure2.png"), combined_plot, 
       width = 7, height = 5, dpi = 500)
```

## 1.6 Outbreaks in Managed Vs. Un-Managed

```{r, eval = T, echo = FALSE}
# Total outbreak over the time
totalOutbreak <- sf::st_read(file.path(analyzed_src, "fireStrata/totalOutbreak.gpkg"))

totalOutbreaks <- totalOutbreak %>% 
  mutate(totalOutbreakA_ha = calculate.area(., "ha")) %>% 
  st_drop_geometry(.) %>% 
  dplyr::select(outbreak_year, outbreak_agents, totalOutbreakA_ha)

# Total outbreak on managed forest over the time
outbreakFuel <- sf::st_read(file.path(analyzed_src, "fireStrata/outbreakFuel.gpkg"))

# Disturbances in Managed forests
managedOutbreak <- outbreakFuel %>% 
  mutate(fuelMgmt = "Managed",
         managedOutbreakA_ha = calculate.area(., "ha")) %>% 
  dplyr::select(outbreak_year, outbreak_agents, fuelMgmt,  managedOutbreakA_ha, geom) %>% 
  st_drop_geometry(.)

managedOutbreak

# Summarize outbreak affected areas
outbreakFuelInfo <- full_join(
  totalOutbreak %>% 
    select(outbreak_year, totalOutbreakA_ha),
  managedOutbreak %>% 
    select(outbreak_year, managedOutbreakA_ha),
  by = "outbreak_year"
) %>%
  # Create long format with managed and unmanaged areas
  mutate(unmanagedOutbreakA_ha = totalOutbreakA_ha - managedOutbreakA_ha) %>%
  pivot_longer(
    cols = c(managedOutbreakA_ha, unmanagedOutbreakA_ha),
    names_to = "mgmtStatus",
    values_to = "OutbreakA_ha"
  ) %>%
  mutate(
    mgmtStatus = case_when(
      mgmtStatus == "managedOutbreakA_ha" ~ "Managed",
      mgmtStatus == "unmanagedOutbreakA_ha" ~ "Unmanaged"
    )
  )

outbreakFuelInfo <-  outbreakFuelInfo %>% 
  mutate(P_OutbreakA_ha = OutbreakA_ha/totalOutbreakA_ha)

outbreakFuelInfo
```

## 1.7 Fire in Managed Vs. Un-Managed

```{r Fire in Managed vs Unmanaged, eval = T, echo = F}

# Total outbreak over the time
totalFire <- sf::st_read(file.path(analyzed_src, "fireStrata/totalFire.gpkg"))
totalFire
totalFireInfo <- totalFire %>% 
  mutate(totalFireA_ha = calculate.area(., "ha"))  %>% 
  st_drop_geometry(.) %>% 
  dplyr::select(fireYear, totalFireA_ha)
totalFireInfo

totalFireA_ha <- sum(totalFireInfo$totalFireA_ha)
totalFireA_ha

# Total outbreak on managed forest over the time
fireFuel <- sf::st_read(file.path(analyzed_src, "fireStrata/fireFuel.gpkg"))

fireFuel
# Disturbances in Managed forests
managedFire <- fireFuel %>% 
  mutate(fuelMgmt = "Managed",
         managedFireA_ha = calculate.area(., "ha")) %>% 
  dplyr::select(fireYear, fuelMgmt,  managedFireA_ha, geom) %>% 
  st_drop_geometry(.)

managedFire

# Summarize outbreak affected areas
fireFuelInfo <- full_join(
  totalFireInfo %>% 
    dplyr::select(fireYear, totalFireA_ha),
  managedFire %>% 
    dplyr::select(fireYear, managedFireA_ha),
  by = "fireYear"
) %>%
  # Create long format with managed and unmanaged areas
  mutate(unmanagedFireA_ha = totalFireA_ha - managedFireA_ha) %>%
  pivot_longer(
    cols = c(managedFireA_ha, unmanagedFireA_ha),
    names_to = "mgmtStatus",
    values_to = "FireA_ha"
  ) %>%
  mutate(
    mgmtStatus = case_when(
      mgmtStatus == "managedFireA_ha" ~ "Managed",
      mgmtStatus == "unmanagedFireA_ha" ~ "Unmanaged"
    )
  )



fireFuelInfo <- fireFuelInfo %>% mutate(P_FireA_ha = FireA_ha/totalFireA_ha)

table(fireFuelInfo$fireYear)
```

## 1.8 Fire in Managed: with vs without outbreak

```{r}
# Fire AND Fuel AND Outbrea, 
fireOutbreakFuel <- sf::st_read(file.path(analyzed_src, "fireStrata/fireOutbreakFuel.gpkg"))
  
fireOutbreakFuel <- fireOutbreakFuel %>%
  mutate(
    outbreak_status = "Outbreak",
    fire_outbreak_area_ha = calculate.area(., "ha")
  ) 


sum(fireOutbreakFuel$fire_outbreak_area_ha)
st_crs(totalOutbreak)
# Get fires without outbreaks using st_difference
fires_no_outbreak <- fireOutbreakFuel %>%
  st_difference(totalOutbreak) %>%
  mutate(
    outbreak_status = "No Outbreak",
    fire_outbreak_area_ha = calculate.area(., "ha")
  ) %>%
  st_drop_geometry(.)

fires_no_outbreak
# Combine the data
fire_outbreak_summary <- bind_rows(
  fireOutbreakFuel,
  fires_no_outbreak
) %>%
  group_by(fire_year, outbreak_status) %>%  # Added fire_year to grouping
  summarize(
    total_area_km2 = sum(fire_outbreak_area_km2),
    .groups = "drop"
  ) %>%
  group_by(fire_year) %>%  # Group by year to calculate percentages within each year
  mutate(
    percent_area = total_area_ha / sum(total_area_ha)
  ) %>%
  ungroup()

fire_outbreak_summary



```

## Figure3

```{r figure3}
# Panel A: Percent area affected by Outbreaks in managed vs unmanaged forests by year
plot_a <- ggplot(outbreakFuelInfo, 
                 aes(x = factor(outbreak_year), y = P_OutbreakA_ha, fill = mgmtStatus)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  labs(x = "Year",
       y = expression("% of total outbreak area per year"),
       fill = "Forest Management",
       title = "Percent area affected by Outbreaks in managed vs unmanaged forests") +
  plot_theme() +
  scale_fill_brewer(palette = "Set2") +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = scales::percent_format())

plot_a

# Panel B: Percent area affected by Fires in managed vs unmanaged forests by year
plot_b <- ggplot(fireFuelInfo, 
                 aes(x = factor(fireYear), y = P_FireA_km2, fill = mgmtStatus)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  labs(x = "Year",
       y = expression("% of total fire area per year"),
       fill = "Forest Management",
       title = "Percent area affected by Fires in managed vs unmanaged forests") +
  plot_theme() +
  scale_fill_brewer(palette = "Set2") +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = scales::percent_format())

plot_b
# Panel C: Fires in managed areas: with vs without subsequent outbreaks by year

plot_c <- fire_outbreak_summary %>%
  ggplot(aes( x = factor(fireYear), y = percent_area,
              fill = outbreak_status)) +
  # Bar plot configuration
  geom_bar(stat = "identity",
           position = "stack",
           width = 0.8) +
  # Color scheme
  scale_fill_manual(
    values = c("No Outbreak" = "#2ca02c","Outbreak" = "#d62728")) +
  # Labels and title
  labs(
    x = "Year",
    y = "% of outbreak area",
    fill = "Outbreak Status",
    title = "(c) Fires in managed areas: with vs without outbreaks"
  ) +
  plot_theme() +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = scales::percent_format())

plot_c

# Combine the plots
combined_plot <- grid.arrange(plot_a, plot_b, plot_c, 
                            ncol = 1, 
                            heights = c(1, 1, 1))

# Save the plot
ggsave(file.path(fig_src, "figure3.png"), combined_plot, 
       width = 7, height = 6, dpi = 300)

```

## Figure4

```{r figure4}

# Panel A: Percent area affected by Outbreaks in managed vs unmanaged forests
plot_a <- ggplot(outbreakFuelInfo, aes(x = factor(outbreak_year), y = outbreakA_km2/sum(outbreakA_km2), 
                                      fill = mgmtStatus)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  labs(x = "Year",
       y = expression("% of total study area"),
       fill = "Management Status",
       title = "Percent area affected by Outbreaks in managed vs unmanaged forests") +
  plot_theme() +
  scale_fill_brewer(palette = "Set2") +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = scales::percent_format())

# Panel B: Percent area affected by Fires in managed vs unmanaged forests
plot_b <- ggplot(fireFuelInfo, aes(x = factor(fireYear), y = FireA_km2/sum(FireA_km2), 
                                  fill = mgmtStatus)) +
  geom_bar(stat = "identity", position = "stack",  width = 0.8) +
  labs(x = "Year",
       y = expression("% of total study area"),
       fill = "Management Status",
       title = "Percent area affected by Fires in managed vs unmanaged forests") +
  plot_theme() +
  scale_fill_brewer(palette = "Set2") +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = scales::percent_format())
   
# Panel C: Fires in managed areas: with vs without subsequent outbreaks
plot_c <- ggplot(fire_outbreak_status %>%
                   group_by(fireYear, outbreak_status) %>%
                   summarize(percent_area = sum(fire_outbreak_area_km2)/sum(fire_outbreak_area_km2)), 
                 aes(x = factor(fire_year), y = percent_area, fill = outbreak_status)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  scale_fill_manual(values = c("No Outbreak" = "#2ca02c", 
                              "Outbreak" = "#d62728"), 
                    labels = c("No Outbreak", "Outbreak")) +
  labs(x = "Year",
       y = expression("% of total study area affected"), 
       fill = "Outbreak Status",
       title = "(c) Fires in managed areas: with vs without outbreaks") +
  plot_theme() +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = scales::percent_format())




# Combine the plots
combined_plot <- grid.arrange(plot_a, plot_b, plot_c, 
                            ncol = 1, 
                            heights = c(1, 1, 1))

# Save the plot
ggsave(file.path(fig_src, "figure4.png"), combined_plot, 
       width = 7, height = 6, dpi = 300)
```

# 2. Sampling Design

## 2.1 Random samples with each stratum

```{r Sampling Design}
# Set to 1GB
options(future.globals.maxSize = 1000 * 1024^2) 
# FIre and Outbreak
fireOutbreak <- sf::st_read(here::here(r"(external\analyzed_data\plots\strata\fireOutbreak.gpkg)"))
fireOutbreak

fireOutbreak <- fireOutbreak %>% filter(FireA_ha > 0)
# fire and fuel
fireFuel <- fireFuel %>% filter(FuelA_ha > 0)
# fire, Outbreak and Fuel
 fireOutbreakFuel <- fireOutbreakFuel %>% filter(OutbreakA_ha > 0)


# Fire Samples
set.seed(345)
fireSamples <- create.samples(fire_lodgepole, areaCol = "FireA_ha", type = "regular", n_cores = 10L)
fireSamples <- do.call(rbind, fireSamples)
future::plan("sequential", .cleanup = T)
gc()

# Fuel Samples
set.seed(456)
fuelSamples <- create.samples(fuel_lodgepole, areaCol = "FuelA_ha", type = "regular", n_cores = 10L)
fuelSamples <- do.call(rbind, fuelSamples)
future::plan("sequential", .cleanup = T)
gc()

# Outbreak
set.seed(567)
outbreakSamples <- create.samples(outbreak_lodgepole, areaCol = "OutbreakA_ha", type = "regular", n_cores = 10L)
outbreakSamples <- do.call(rbind, outbreakSamples)
future::plan("sequential", .cleanup = T)
gc()


fireSamples
fuelSamples
outbreakSamples

outbreak_lodgepole

st_write(fireSamples, file.path(analyzed_src, "plots/strata/fireSamples.gpkg"))
st_write(fuelSamples, file.path(analyzed_src, "plots/strata/fuelSamples.gpkg"))
st_write(outbreakSamples, file.path(analyzed_src, "plots/strata/outbreakSamples.gpkg"))
```

## 2.2 Load samples from all strata

```{r}
allStrataWithFire <- sf::st_read(file.path(analyzed_src, "plots/strata/allStrataWithFire.gpkg"))




```

## 2.3 Combine and Label samples with each stratum

```{r}
stratifiedSamples <- allStrataWithFire %>%
  mutate(
    in_fire = if_else(!is.na(fire_id_2), 1, 0),
    in_fuel = if_else(!is.na(fuel_year), 1, 0),
    in_outbreak = if_else(!is.na(outbreak_year), 1, 0),

    # Create strata label based on intersections
    strata = case_when(
      in_fire == 1 & in_fuel == 1 & in_outbreak == 1 ~ "fire_fuel_outbreak",
      in_fire == 1 & in_fuel == 1 ~ "fire_fuel",
      in_fire == 1 & in_outbreak == 1 ~ "fire_outbreak",
      in_fire == 1 ~ "fire_only",
      TRUE ~ "none"  # points that don't fall in any polygon
    )
  )

table(stratifiedSamples$strata)
```

## 2.4 Random Non-fire samples

```{r nonFireSamples}
# Create allFireA
allFireA <- st_union(fire_lodgepole)

# Create the non-Fire area within Lodgepole pine forests area.
nonFireA <- st_difference(study_area, allFireA)

st_write(nonFireA, file.path(analyzed_src, 
                                "mtbs_fire_data/nonFireA.gpkg"), append = F)
# ====================================================================================
# Load nonFireA.gpkg
nonFireA <- st_read(file.path(analyzed_src, 
                                "mtbs_fire_data/nonFireA.gpkg"))


# Fire date count to make a pair with nonFire points
fire_date_counts <- fireSamples %>%
  group_by(fire_date) %>%
  summarise(n = n()) %>%
  arrange(fire_date)


# Create point for nonFireArea
set.seed(123)
nonFireSamples <- sf::st_sample(nonFireA, 
                                size = nrow(fireSamples) * 2, 
                                type = "regular") %>% st_as_sf(.) %>% 
  
  dplyr::filter(!rowSums(st_intersects(., allFireA, sparse = F)) > 0) %>% 
  slice_sample(n = nrow(fireSamples))


# Randomly assign fire dates to nonFireSamples
set.seed(123)  
nonFireSamples<- nonFireSamples %>%
  mutate(fire_date = sample(
    rep(fire_date_counts$fire_date, fire_date_counts$n),
    size = n(),
    replace = FALSE
  )) %>% 
  mutate(plot_label = "nonFire",
         fire_id = "non_fire_event",
         fireYr = lubridate::year(fire_date)) %>% 
  rename(geom = x)

nonFireSamples
  
st_write(nonFireSamples, file.path(analyzed_src, 
                                "mtbs_fire_data/nonFireSamples.gpkg"), append = F)

```

## 2.5 fireNonFire Plots

```{r}
# fire samples
fireSamples <- sf::st_read(file.path(analyzed_src, 
                                "mtbs_fire_data/fireSamples.gpkg"))
# nonFire samples
nonFireSamples <- sf::st_read(file.path(analyzed_src, 
                                "mtbs_fire_data/nonFireSamples.gpkg"))
fireSamples;nonFireSamples


# Bind fire and nonFire samples by row 
fireNonFirePlots <- bind_rows(fireSamples, nonFireSamples) %>% 
  mutate(plotId = paste0(plot_label,1:nrow(.))) %>%
  dplyr::rename(plotLabel = plot_label,
                fireDate = fire_date) %>% 
  dplyr::select(fire_id, plotId, plotLabel, fireDate, fireYr, geom)

table(fireNonFirePlots$plotLabel)
# ====================================================================================

```

## 2.6 Randomly selecting samples from the lists

```{r samplePlots}
samplePlots <- fireNonFirePlots %>% 
  group_by(fireYr, plotLabel) %>%
  slice_sample(n = 30) %>%
  ungroup()


# Write final samplePlots to the file 
sf::st_write(samplePlots, 
             file.path(analyzed_src, "plots/samplePlots.gpkg"),
             append = F
             )
samplePlots
```

## 2.7 Load Final samplePlots

```{r fireNonFirePlots}
samplePlots <- sf::st_read(file.path(analyzed_src, filename, "samples.gpkg"))

```

# 3. Climate Variables

## 3.1 Monthly Mean Maximum Temperature

### 3.1.1 Calculate monthly tmmx

```{r}
var_name <- getGridmet.metadata() %>% 
  dplyr::filter(variable == "tmmx") %>% 
  pull(Description)
var_name  

fileSrc <- here::here("external/raw_data/climate_data", "tmmx")
fileList <- dir(fileSrc, pattern = "nc", full.names = T)
outputDir <- here::here("external/analyzed_data/climate_data", "tmmx", "monthly")

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

# ===========================================================================
future::plan("multisession", workers = 4L)

# Create progress handler
progressr::handlers(global = T)
progressr::handlers("progress")
progressr::with_progress({
  p <- progressr::progressor(along = fileList)
  
  result <- future.apply::future_lapply(fileList, function(fl) {
     
        gridmetDaily2Monthly(fl, var_name, outputDir) 
      
        p()
     })
  })
future::plan("sequential", .cleanup = T)


```

### 3.1.2 Extract monthly tmmx values to point

```{r tmmx monthly}

tmmxM_dir <- here::here("external/analyzed_data/climate_data", "tmmx", "monthly")

samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                           rastDir = tmmxM_dir,
                                           variable = "tmmxM",
                                           match.column = "fireDate"
                                           )

samplePlots <- samplePlots %>% mutate(tmmxM = tmmxM - 273.15)
st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)

```

## 3.2 Monthly Mean Minimum Temperature

### 3.2.1 Calculate monthly tmmn

```{r tmmn monthly}
var_name <- getGridmet.metadata() %>% 
  dplyr::filter(variable == "tmmn") %>% 
  pull(Description)
var_name  

fileSrc <- here::here("external/raw_data/climate_data", "tmmn")
fileList <- dir(fileSrc, pattern = "nc", full.names = T)
outputDir <- here::here("external/analyzed_data/climate_data", "tmmn", "monthly")

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

# ===========================================================================
future::plan("multisession", workers = 4L)

# Create progress handler
progressr::handlers(global = T)
progressr::handlers("progress")
progressr::with_progress({
  p <- progressr::progressor(along = fileList)
  
  result <- future.apply::future_lapply(fileList, function(fl) {
     
        gridmetDaily2Monthly(fl, var_name, outputDir) 
      
        p()
     })
  })
future::plan("sequential", .cleanup = T)
```

### 3.2.2 Extract monthly tmmn values to points

```{r}
tmmnM_dir <- here::here("external/analyzed_data/climate_data", "tmmn", "monthly")

samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                           rastDir = tmmnM_dir,
                                           variable = "tmmnM",
                                           match.column = "fireDate"
                                           )
samplePlots <- samplePlots %>% mutate(tmmnM = tmmnM - 273.15)
samplePlots

st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)


```

## 3.3 Monthly Mean Temperature

### 3.3.1 Calculate monthly tmean

```{r tmean monthly}
tmmnM_dir <- here::here("external/analyzed_data/climate_data", "tmmn", "monthly")
tmmxM_dir <- here::here("external/analyzed_data/climate_data", "tmmx", "monthly")
tmmnMList <- dir(tmmnM_dir, pattern = "nc", full.names = T)
tmmxMList <- dir(tmmxM_dir, pattern = "nc", full.names = T)

outputDir <- here::here("external/analyzed_data/climate_data", "tmean", "monthly")

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

tmeanM <-  calculate_tmeanM(tmmnMList, tmmxMList, outputDir = outputDir)


# ===========================================================================


```

### 3.3.2 Extract monthly tmean values to points

```{r}
tmeanM_dir <- here::here("external/analyzed_data/climate_data", "tmean", "monthly")

samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                           rastDir = tmeanM,
                                           variable = "tmeanM",
                                           match.column = "fireDate"
                                           )

samplePlots
st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)
```

## 3.4 Precipitation

### 3.4.1 Calculate monthly pr

```{r Precipitation}
var_name <- getGridmet.metadata() %>% 
  dplyr::filter(variable == "pr") %>% 
  pull(Description)
var_name  

fileSrc <- here::here("external/raw_data/climate_data", "pr")
fileList <- dir(fileSrc, pattern = "nc", full.names = T)
outputDir <- here::here("external/analyzed_data/climate_data", "pr", "monthly")

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

ncdf4::nc_open(fileList[1])
# ===========================================================================
future::plan("multisession", workers = 4L)

# Create progress handler
progressr::handlers(global = T)
progressr::handlers("progress")
progressr::with_progress({
  p <- progressr::progressor(along = fileList)
  
  result <- future.apply::future_lapply(fileList, function(fl) {
     
        gridmetDaily2Monthly(fl, var_name, outputDir) 
      
        p()
     })
  })
future::plan("sequential", .cleanup = T)
```

### 3.4.2 Extract monthly pr values to points

```{r}
prM_dir <- here::here("external/analyzed_data/climate_data", "pr", "monthly")

samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                           rastDir = prM_dir,
                                           variable = "prM",
                                           match.column = "fireDate"
                                           )

samplePlots
st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)


        
```

## 3.5 PDSI Monthly

### 3.5.1 Download pdsi monthly

```{r}
basePath <- file.path(here::here("external/raw_data/climate_data"))
pdsiM <- getPrismDataset("pdsi", 1984, 2023, basePath)

```

### 3.5.2 Extract monthly pdsi values to points

```{r}
pdsiM_dir <- here::here("external/raw_data/climate_data", "pdsi")


samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                           rastDir = pdsiM_dir,
                                           variable = "pdsiM",
                                           match.column = "fireDate"
                                           )


st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)


```

## 3.6 SC-PDSI Monthly

### 3.6.1 Download scPdsi monthly

```{r}
basePath <- file.path(here::here("external/raw_data/climate_data"))
scPdsiM <- getPrismDataset("scpdsi", 1984, 2023, basePath)
```

### 3.6.2 Extract monthly scPdsi values to points

```{r}
scPdsiM_dir <- here::here("external/raw_data/climate_data", "scpdsi")


samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                           rastDir = scPdsiM_dir,
                                           variable = "scPdsiM",
                                           match.column = "fireDate"
                                           )


samplePlots
st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)
```

## 3.7 PZi Monthly

### 3.7.1 Download pzi monthly

```{r}
basePath <- file.path(here::here("external/raw_data/climate_data"))
pziM <- getPrismDataset("pzi", 1984, 2023, basePath)
```

### 3.7.2 Extract monthly pzi values to points

```{r}
pzIM_dir <- here::here("external/raw_data/climate_data", "pzi")


samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                           rastDir =pzIM_dir,
                                           variable = "pzIM",
                                           match.column = "fireDate"
                                           )


samplePlots
st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)
```

## 3.8 vpd Monthly

### 3.8.1 Calculate Monthly vpd from Daily vpd

```{r}

var_name <- getGridmet.metadata() %>% 
  dplyr::filter(variable == "vpd") %>% 
  pull(Description)
  
fileSrc <- here::here("external/raw_data/climate_data", "vpd")
fileList <- dir(fileSrc, pattern = "nc", full.names = T)
outputDir <- here::here("external/analyzed_data/climate_data", "vpd/monthly")

# ---------------------------------------------------------------------
future::plan("multisession", workers = 4L)

# Create progress handler
progressr::handlers(global = T)
progressr::handlers("progress")
progressr::with_progress({
  p <- progressr::progressor(along = fileList)
  
  result <- future.apply::future_lapply(fileList, function(fl) {
     
        gridmetDaily2Monthly(fl, var_name, outputDir) 
      
        p()
     })
  })
future::plan("sequential", .cleanup = T)
```

### 3.8.2 Extract vpd value to the points

```{r}
vpdM_dir <- here("external/analyzed_data/climate_data/vpd/monthly")

samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                           rastDir = vpdM_dir,
                                           variable = "vpdM",
                                           match.column = "fireDate"
                                           )
st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)
```

## 3.9 pet Monthly

### 3.9.1 Calculate pet Monthly from daily data

```{r}

var_name <- getGridmet.metadata() %>% 
  dplyr::filter(variable == "pet") %>% 
  pull(Description)
  
fileSrc <- here::here("external/raw_data/climate_data", "pet")
fileList <- dir(fileSrc, pattern = "nc", full.names = T)
outputDir <- here::here("external/analyzed_data/climate_data", "pet", "monthly")
if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

# ---------------------------------------------------------------------
future::plan("multisession", workers = 4L)

# Create progress handler
progressr::handlers(global = T)
progressr::handlers("progress")
progressr::with_progress({
  p <- progressr::progressor(along = fileList)
  
  result <- future.apply::future_lapply(fileList, function(fl) {
     
        gridmetDaily2Monthly(fl, var_name, outputDir) 
      
        p()
     })
  })
future::plan("sequential", .cleanup = T)

```

### 3.9.2 Extract pet monthly values by point

```{r}
petM_dir <- here("external/analyzed_data/climate_data/pet/monthly")

samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                           rastDir = petM_dir,
                                           variable = "petM",
                                           match.column = "fireDate"
                                           )

st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)
```

## 3.10 vs Monthly

### 3.10.1 Calculate vs Monthly from vs Daily

```{r}
var_name <- getGridmet.metadata() %>% 
  dplyr::filter(variable == "vs") %>% 
  pull(Description)
  
fileSrc <- here::here("external/raw_data/climate_data", "vs")
fileList <- dir(fileSrc, pattern = "nc", full.names = T)
outputDir <- here::here("external/analyzed_data/climate_data", "vs", "monthly")
if(!dir.exists(outputDir)){
  dir.create(outputDir)
}


# ---------------------------------------------------------------------
future::plan("multisession", workers = 4L)

# Create progress handler
progressr::handlers(global = T)
progressr::handlers("progress")
progressr::with_progress({
  p <- progressr::progressor(along = fileList)
  
  result <- future.apply::future_lapply(fileList, function(fl) {
     
        gridmetDaily2Monthly(fl, var_name, outputDir) 
      
        p()
     })
  })
future::plan("sequential", .cleanup = T)
```

### 3.10.2 Extract

```{r}
vsM_dir <- here("external/analyzed_data/climate_data/vs/monthly")

samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                           rastDir = vsM_dir,
                                           variable = "vsM",
                                           match.column = "fireDate"
                                           )
st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)
```

## 3.11 rmax monthly

### 3.11.1 Calculate monthly rmax

```{r}
var_name <- getGridmet.metadata() %>% 
  dplyr::filter(variable == "rmax") %>% 
  pull(Description)
  
fileSrc <- here::here("external/raw_data/climate_data", "rmax")
fileList <- dir(fileSrc, pattern = "nc", full.names = T)
outputDir <- here::here("external/analyzed_data/climate_data", "rmax", "monthly")

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

# ---------------------------------------------------------------------
future::plan("multisession", workers = 4L)

# Create progress handler
progressr::handlers(global = T)
progressr::handlers("progress")
progressr::with_progress({
  p <- progressr::progressor(along = fileList)
  
  result <- future.apply::future_lapply(fileList, function(fl) {
     
        gridmetDaily2Monthly(fl, var_name, outputDir) 
      
        p()
     })
  })
future::plan("sequential", .cleanup = T)

```

### 3.11.2 Extract rmax monthly values by point

```{r}
rmaxM_dir <- here("external/analyzed_data/climate_data/rmax/monthly")

samplePlots <- extractValue2Points(samplePlots, 
                                           rastDir = rmaxM_dir,
                                           variable = "rmaxM",
                                           match.column = "fireDate"
                                           )
      

st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)
samplePlots
```

## 3.12 rmin Monthly

```{r}
var_name <- getGridmet.metadata() %>% 
  dplyr::filter(variable == "rmin") %>% 
  pull(Description)
var_name  
fileSrc <- here::here("external/raw_data/climate_data", "rmin")
fileList <- dir(fileSrc, pattern = "nc", full.names = T)
outputDir <- here::here("external/analyzed_data/climate_data", "rmin", "monthly")

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}


# ---------------------------------------------------------------------
future::plan("multisession", workers = 4L)

# Create progress handler
progressr::handlers(global = T)
progressr::handlers("progress")
progressr::with_progress({
  p <- progressr::progressor(along = fileList)
  
  result <- future.apply::future_lapply(fileList, function(fl) {
     
        gridmetDaily2Monthly(fl, var_name, outputDir) 
      
        p()
     })
  })
future::plan("sequential", .cleanup = T)


```

### 3.12.2 Extract rmin monthly values to point

```{r}
rminM_dir <- here("external/analyzed_data/climate_data/rmin/monthly")

samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                           rastDir = rminM_dir,
                                           variable = "rminM",
                                           match.column = "fireDate"
                                           )


st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)
```

## 3.13 srad Monthly

### 3.13.1 Calculate srad Monthly from daily srad data

```{r}
var_name <- getGridmet.metadata() %>% 
  dplyr::filter(variable == "srad") %>% 
  pull(Description)
var_name  
fileSrc <- here::here("external/raw_data/climate_data", "srad")
fileList <- dir(fileSrc, pattern = "nc", full.names = T)
outputDir <- here::here("external/analyzed_data/climate_data", "srad", "monthly")

if(!dir.exists(outputDir)){
  dir.create(outputDir)
}

# ===========================================================================
future::plan("multisession", workers = 4L)

# Create progress handler
progressr::handlers(global = T)
progressr::handlers("progress")
progressr::with_progress({
  p <- progressr::progressor(along = fileList)
  
  result <- future.apply::future_lapply(fileList, function(fl) {
     
        gridmetDaily2Monthly(fl, var_name, outputDir) 
      
        p()
     })
  })
future::plan("sequential", .cleanup = T)

```

### 3.13.2 Extract srad monthly values to points

```{r}
sradM_dir <- here("external/analyzed_data/climate_data/srad/monthly")

samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                           rastDir = sradM_dir,
                                           variable = "sradM",
                                           match.column = "fireDate"
                                           )

samplePlots
st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)
```

# 4. Topographic Variables

## 4.1 Elevation

```{r}
# DEM
dem_30m <- file.path(raw_src,"topo_data/LF2020_Elev_220_CONUS/LF2020_Elev_220_CONUS/Tif/LC20_Elev_220.tif")

samplePlots

# Extract Elevation Data
samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                    rastDir = dem_30m, 
                                    variable = "elev30M", match.column = F)


st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)
```

## 4.2 Slope

```{r}
#  Slope
slope_30M <- file.path(raw_src,"topo_data/LF2020_SlpD_220_CONUS/LF2020_SlpD_220_CONUS/Tif/LC20_SlpD_220.tif")

# Extract Slope Data
samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                    rastDir = slope_30M, 
                                    variable = "slpD30M", match.column = F)


st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)
```

## 4.3 Aspect

```{r}
# Aspect
aspect_30m <- file.path(raw_src,"topo_data/LF2020_Asp_220_CONUS/LF2020_Asp_220_CONUS/Tif/LC20_Asp_220.tif")

# Extract Aspect Data
samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                    rastDir = aspect_30m, 
                                    variable = "aspD30M", match.column = F)


st_write(samplePlots, 
        file.path(plots_src, "samplePlots.gpkg"), append = FALSE)
samplePlots
```

## 4.4 TPI

```{r}
tpi_conus <- file.path(analyzed_src,"topographic/tpi30m.tif")


# Extract TPI Data
samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                    rastDir = tpi_conus, 
                                    variable = "tpi", match.column = F)

st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)
table(is.na(samplePlots$tpi))
```

## 4.5 TWI

```{r}
twi_conus <- file.path(analyzed_src,"topographic/twi_30m.tif")
file.exists(twi_conus)
samplePlots <- extractValue2Points(pointFeature = samplePlots, 
                                    rastDir = twi_conus, 
                                    variable = "twi", match.column = F)



st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)

```

# 5. Burn Severity

The severity classes are:

0: Background;

1: Unburned to Low; 2: Low; 3: Moderate; 4: High; 5: Increased Greenness;

6: Non-Processing Area Mask

## 5.1 Extract all TIF files into one directory
```{r}

# Set paths
source_dir <- here("external/raw_data/mtbs_fire_data/Fire_data_bundles/composite_data/MTBS_BSmosaics")
output_dir <- here("external/raw_data/mtbs_fire_data/burnSeverity_tifs")  

# Create output directory if it doesn't exist
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# List all zip files recursively
zip_files <- list.files(path = source_dir, 
                       pattern = "mtbs_CONUS_.*\\.zip$", 
                       recursive = TRUE,
                       full.names = TRUE)

# Function to extract tif from zip
extract_tif <- function(zip_path, out_dir) {
  tryCatch({
    # Create a temporary directory
    temp_dir <- tempdir()
    
    # Unzip to temp directory
    unzip(zip_path, exdir = temp_dir)
    
    # Find the .tif file
    tif_file <- list.files(temp_dir, 
                          pattern = "\\.tif$", 
                          recursive = TRUE, 
                          full.names = TRUE)
    
    # Move tif file to output directory
    if(length(tif_file) > 0) {
      file.copy(tif_file, 
                file.path(out_dir, basename(tif_file)), 
                overwrite = TRUE)
    }
    
    # Clean up temp directory
    unlink(temp_dir, recursive = TRUE)
  }, error = function(e) {
    message("Error processing ", zip_path, ": ", e$message)
  })
}

# Process all zip files
for(zip_file in zip_files) {
  message("Processing: ", basename(zip_file))
  extract_tif(zip_file, output_dir)
}


```

## 5.2 Masking and Reprojecting burnSeverity
```{r}
# Masking 
rawMtbsI_src <- dir(file.path("external/raw_data/mtbs_fire_data/burnSeverity_tifs"), pattern = ".tif", full.names = T)
output_src <- file.path(analyzed_src, "mtbs_fire_data/mtbs_severity")

# Transform to crs of raster data 
vecConusStates <- st_transform(conusStates, st_crs(rast(rawMtbsI_src[40])))

# clip burn severity
raster_clip(rawMtbsI_src[40], vecConusStates, output_src)

# Western CONUS burn severity raster data
conusMtbsI_src <- dir(file.path(analyzed_src, "mtbs_fire_data/mtbs_severity"), pattern = ".tif", full.names = T)

# Reproject fire severity raster for 2023
mtbs_CONUS_2023 <- project(rast(conusMtbsI_src[40], "EPSG:5070"))

# Write a raster to the file
writeRaster(mtbs_CONUS_2023, file.path(conusMtbsI_src, "conus_epsg5070", "mtbs_CONUS_2023.tif"))






```


# 6. Edaphic factors

## 6.1 Available water capacity

```{r}
# Load awc dataset
awcDir <- dir(here::here("external/raw_data/topo_data/awc_gNATSGO/awc_gNATSGO"),
              pattern = ".tif$", full.names = T)

# Extract the data to the points
samplePlots <- extractValue2Points(samplePlots, awcDir, variable = "awC")

# Write a samplePlots object to the file
st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)

```

## 6.2 Field Capacity

```{r}
# Load fc dataset
fcDir <- dir(here::here("external/raw_data/topo_data/fc_gNATSGO/fc_gNATSGO"),
              pattern = ".tif$", full.names = T)

# Extract the data to the points
samplePlots <- extractValue2Points(samplePlots, fcDir, variable = "fC")

# Write a samplePlots object to the file
st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)
```

## 6.3 Soil porosity

```{r}
# Load awc dataset
porDir <- dir(here::here("external/raw_data/topo_data/por_gNATSGO/por_gNATSGO"),
              pattern = ".tif$", full.names = T)

# Extract the data to the points
samplePlots <- extractValue2Points(samplePlots, porDir, variable = "por")

# Write a samplePlots object to the file
st_write(samplePlots, 
         file.path(plots_src, "samplePlots.gpkg"),
         append = F)

samplePlots
```

# 7. Tree Canopy Cover

The USFS Forest Inventory and Analysis (FIA) program photo-interpreted percent tree canopy cover (TCC) response data. Photointerpretation (PI) measured TCC using a custom ArcGIS plug-in tool (Goeking et al., 2012) from 105-point grids placed in 90x90 squares centered on USFS FIA plot design. A total of 55,242 PI plots were used in TCC modeling.

```         
https://doi.org/10.1016/j.isprsjprs.2018.09.006
```

```{r}
# Load PTCC dataset
treePcc_Src <- dir(here::here("external/raw_data/treecanopy_database"), 
                  pattern = ".tif$", full.names = T)

# Extract the data to the points
samplePlots <- extractValue2Points(samplePlots, treePcc_Src, variable = "treePcc")


# Write a samplePlots object to the file
st_write(samplePlots, 
         file.path(aplots_src, "samplePlots.gpkg"),
         append = F)
```

# 8. Statistical Analysis

## 8.1 Disturbance Classes

```{r}
outbreak_lodgepole

```

# 9. Modeling

## 9.1 Load samplePlots

```{r}
# Load fire NonFire plots as samplePlots 
samplePlots <- st_read(here::here("external/analyzed_data/plots/samplePlots.gpkg")) 

samplePlots <- samplePlots %>% mutate(fireYrs = lubridate::year(fireDate))
samplePlots

# Total number of observations is 77062.
samplePlots <- samplePlots %>%
  mutate(FNF = case_when(
           plotLabel == "fire" ~ "1",
           plotLabel == "nonfire" ~ "0"
         )) %>% 
  create.timeclass(., "fireYear", 1984, 2023, 10, "fire") %>% 
  st_transform("epsg:5070")

```

## 9.2 Outbreak and Fuel Info

```{R}        
samplePlots <- samplePlots %>%  mutate(
    fuel = as.numeric(strata == "fireFuel"),
    outbreak = as.numeric(strata == "fireOutbreak"),
    outbreakFuel = as.numeric(strata == "fireOutbreakFuel")
  )

samplePlots1 <- samplePlots %>% 
  # Update outbreak based on intersection
  mutate(outbreak =  if_else(
    outbreak == 0 & lengths(st_intersects(., outbreakLodgepole)) > 0, 1, 0)
    )
  samplePlots1 %>% filter(plotLabel == "nonFire" & outbreak == 1)
  table(samplePlots$outbreak)
  # )) %>%
  # # Update fuel based on intersection
  # mutate(fuel = case_when(
  #   to_update & lengths(st_intersects(., fuelLodgepole)) > 0 ~ 1,
  #   TRUE ~ fuel
  # )) %>%
  # # Remove the temporary flag
  # select(-to_update)



st_crs(outbreakLodgepole)
st_crs(samplePlots)

samplePlots

# Write a samplePlots object to the file
st_write(samplePlots, 
         file.path(analyzed_src, "plots/samplePlots.gpkg"),
         append = F)
```


## 9.4 Pre-processing the dataset

## 9.5 Selecting predictors

```{r Spliting dataset}
# All numerical features from samplePlots
corrCol <- c("mtbsI", "tmmxM", "tmmnM", "prM", "pdsiM",  "scPdsiM", "pzIM", "vpdM",  
"petM", "vsM", "rmaxM", "rminM", "sradM", "elev30M", "slpD30M", "aspD30M",  "awC", "tpi", "fuel", "outbreak", "outbreakFuel")

# Coorelation plot
corrPlot(samplePlots, corrCol)
colnames(samplePlots)

# Target variable for fire and fire severity
facCol <- c("plotLabel", "severity")
dependentVar <- c("FNF", "mtbsI")

numCol <- c( "tmmxM", "prM", "pdsiM",  "vpdM", 
"petM", "vsM", "sradM", "elev30M", 
"slpD30M", "aspD30M",  "awC", "tpi")

colList <- c(facCol, numCol)
colList

table(samplePlots$strata)
cor_matrix <- cor(samplePlots %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(all_of(corrCol)) %>% 
  mutate(across(all_of(numCol), as.numeric)) %>% 
  as.data.table())

cor_matrix
high_cor <- findCorrelation(cor_matrix, cutoff = 0.75)
X_filtered <- X[, -high_cor]

```

### 9.5.1 Splitting the dataset for training and testing Models

```{r Final Dataset}
data <- samplePlots %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(all_of(colList)) %>% 
  mutate(across(all_of(facCol), as.factor)) %>% 
  mutate(across(all_of(numCol), as.numeric)) %>% 
  as.data.table()  %>% 
  .[, (numCol) := lapply(.SD, nafill, type = "locf"), .SDcols = numCol]


  
# Dataset for fire nonfire
data_FNF <- data %>% 
  dplyr::select(plotLabel, all_of(numCol))
str(data_FNF)

# write.csv(data_FNF, here::here("data/samples/data_FNF.csv"))

# Dataset for mtbsI
data_mtbsI <- data %>% 
  dplyr::select(severity, all_of(numCol), -plotLabel)




str(data_mtbsI)
# write.csv(data_mtbsI, here::here("data/samples/data_mtbsI.csv"))

```

### 9.5.2 Spliting the datasets

```{r}

# Dataset summary
nrow(data_FNF); table(data_FNF$plotLabel)
nrow(data_mtbsI); table(data_mtbsI$mtbsI)

# Split the data into training and testing sets
set.seed(246)
sample_FNF <- sample.int(n = nrow(data_FNF), size = floor(.60*nrow(data_FNF)), replace = F)
sample_mtbsi <- sample.int(n = nrow(data_mtbsI), size = floor(.60*nrow(data_mtbsI)), replace = F)

train_data_FNF <- data_FNF[sample_FNF, ]
test_data_FNF  <- data_FNF[-sample_FNF,]

train_data_mtbsi <- data_mtbsI[sample_mtbsi, ]
test_data_mtbsi  <- data_mtbsI[-sample_mtbsi, ]
str(train_data_FNF)
str(train_data_mtbsi)
```

### 9.5.3 Handling predictors and response for XGBoost model

```{r}
missing_pred <- colSums(is.na(data_FNF[, ..numCol]))
missing_pred
  
missing_target <- sum(is.na(data_FNF[["plotLabel"]]))
missing_target


missing_pred <- colSums(is.na(data_mtbsI[, ..numCol]))
missing_pred
  
missing_target <- sum(is.na(data_mtbsI[["severity"]]))
missing_target

```

## 9.6 RandomForest Model

```{r}
library(randomForest)
library(caret)
library(doParallel)

# Setup parallel processing
registerDoParallel(cores = 4)

# Create formula
formFire <- as.formula(paste("event", "~", paste(numCol,  collapse = " + ")))

# Set up train control for repeated CV
ctrl <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 5,
  allowParallel = TRUE,
  savePredictions = "final",
  returnResamp = "all"
)

# Train Random Forest using caret for CV
rf_cv <- train(
  formFire,
  data = train_data_FNF,
  method = "rf",
  trControl = ctrl,
  ntree = 2000,
  importance = TRUE
)

print(rf_cv)

# ====================================================================================
# Fire Severity
# ====================================================================================
formSeverity <- as.formula(paste("severity", "~", paste(numCol, collapse = " + ")))
# Train Random Forest using caret for CV
rf_cv_mtbsi <- train(
  formSeverity,
  data = train_data_mtbsi,
  method = "rf",
  trControl = ctrl,
  ntree = 2000,
  importance = TRUE
)



print(rf_model_mtbsi)

# ====================================================================================
caret::varImp(rf_cv)

# Make predictions on the test dataset
pred_model_FNF <- predict(rf_cv, newdata = test_data_FNF[,-1])
pred_model_severity <-  predict(rf_cv_mtbsi, newdata = test_data_mtbsi[,-1])



# Calculate the confusion matrix
conf_matrix_FNF <- confusionMatrix(pred_model_FNF, as.factor(test_data_FNF$plotLabel))
conf_matrix_FNF
conf_matrix_mtbsi <- confusionMatrix(pred_model_severity, as.factor(test_data_mtbsi$severity))
conf_matrix_mtbsi

# Plot the Random Forest model performance
plot(rf_cv, main = "Model performance of RandomForest model for fire")

plot(rf_cv_mtbsi, main = "Model performance of RandomForest model for fire severity")



# Plot the importance of variables
important_variables <- varImp(rf_cv, scale = T)
plot_a <- plot(important_variables, 
     type = "p", 
     main = "Feature importance plot using RF model for predicting Fire",
     ylab = "Features")


important_variables_mtbsi <- varImp(rf_cv_mtbsi, scale = T)
plot_b <- plot(important_variables_mtbsi, 
     type = "p", 
     main = "Feature importance plot using RF model for predicting Fire Severity",
     ylab = "Features")

# ====================================================================================

plot_a
# Save the plot
ggsave(file.path(fig_src, "figure5.png"), plot_a, 
       width = 7, height = 6, dpi = 300)

ggsave(file.path(fig_src, "figure6.png"), plot_b, 
       width = 7, height = 6, dpi = 300)
```

## MDS plot

```{r}
# Get proximity matrix
prox_matrix <- rf_model$proximity

# Convert proximity to distance
dist_matrix <- sqrt(1 - prox_matrix)

# Perform MDS on distance matrix
mds <- cmdscale(dist_matrix, k = 2)

# Create visualization dataset
viz_data <- data.frame(
  MDS1 = mds[,1],
  MDS2 = mds[,2],
  Class = train_data_FNF[[target_var]]
)

# Create separability plot
sep_plot <- ggplot(viz_data, aes(x = MDS1, y = MDS2, color = Class)) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Random Forest Proximity-based Class Separability",
    x = "MDS Dimension 1",
    y = "MDS Dimension 2"
  )

# Calculate CV performance metrics
cv_results <- rf_cv$results
cv_resamples <- rf_cv$resample

# Calculate separability metrics
within_class_dist <- list()
between_class_dist <- list()
classes <- unique(data[[target_var]])

for(class in classes) {
  class_indices <- which(train_data_FNF[[target_var]] == class)
  within_class_dist[[as.character(class)]] <- mean(dist_matrix[class_indices, class_indices])
  
  other_indices <- which(train_data_FNF[[target_var]] != class)
  between_class_dist[[as.character(class)]] <- mean(dist_matrix[class_indices, other_indices])
}

# Calculate average separability ratio
separability_ratio <- mean(unlist(between_class_dist)) / mean(unlist(within_class_dist))

# Return results
list(
  cv_model = rf_cv,
  final_model = rf_model,
  cv_results = cv_results,
  cv_resamples = cv_resamples,
  proximity_matrix = prox_matrix,
  distance_matrix = dist_matrix,
  mds_coords = mds,
  separability_plot = sep_plot,
  within_class_distances = within_class_dist,
  between_class_distances = between_class_dist,
  separability_ratio = separability_ratio
)
```

## Modelling

```{r}

```

### 9.6.1 Fire occurrence as response variable

```{r Random Forest}
library(doParallel)
library(caret)

cl <- makeCluster(4)
registerDoParallel(cl)


# Define the control for cross-validation
train_control <- trainControl(method = "repeatedcv", 
                              number = 10, 
                              repeats = 5,
                              allowParallel = T,
                              )


mtry_grid_FNF <- expand.grid(mtry = c(2, 3, 4, 5, 7, 8))


# Train the random forest model
set.seed(123)
fnfModel_RF <- train(plotLabel ~ .,
                            data = train_data_FNF, 
                            method = "rf", 
                            trControl = train_control,
                            tuneGrid = mtry_grid_FNF,
                            ntree = 2000,
                            importance = T,
                     proximity = T
                            
                            )

stopCluster(cl)



# Save RF_model
saveRDS(fnfModel_RF, file.path(analyzed_src, "RandomForest/fnfModel_RF_1.rds"))
saveRDS(fnfModel_RF, file.path(analyzed_src, "RandomForest/fnfModel_RF_2.rds"))
saveRDS(fnfModel_RF, file.path(analyzed_src, "RandomForest/fnfModel_RF_3.rds"))
saveRDS(fnfModel_RF, file.path(analyzed_src, "RandomForest/fnfModel_RF_4.rds"))
saveRDS(fnfModel_RF, file.path(analyzed_src, "RandomForest/fnfModel_RF_5.rds"))
# Print model
print(fnfModel_RF)

# Load save model
fnfModel_RF <- readRDS(file.path(here("external/analyzed_data/RandomForest/fnfModel_RF_3.rds")))

# Make predictions on the test dataset
pred_model_FNF <- predict(fnfModel_RF, newdata = test_data_FNF[,-1])

# Calculate the confusion matrix
conf_matrix_FNF <- confusionMatrix(pred_model_FNF, as.factor(test_data_FNF$plotLabel))
conf_matrix_FNF

# Plot the Random Forest model performance
plot(fnfModel_RF, main = "Model performance of RandomForest model for fire")


# Plot the importance of variables
important_variables <- varImp(fnfModel_RF, scale = T)
plot(important_variables, 
     type = "p", 
     main = "Feature importance plot using RF model for predicting Fire",
     ylab = "Features")

```

### 9.6.2 Fire severity as the response variable

```{r}
cl <- makeCluster(4)
registerDoParallel(cl)

mtry_grid_mtbsi <- expand.grid(mtry = c(2, 3, 4, 5, 7, 8))
 

# Define the control for cross-validation
train_control <- trainControl(method = "repeatedcv", 
                              number = 10, 
                              repeats = 5, 
                              allowParallel = T)

set.seed(589)
mtbsiModel_RF <- train(severity ~ .,
                   data = train_data_mtbsi, 
                   method = "rf", 
                   trControl = train_control,
                   tuneGrid = mtry_grid_mtbsi,
                   ntree = 2000,
                   importance = T)

stopCluster(cl)

# Save RF_model
saveRDS(mtbsiModel_RF, file.path(analyzed_src, "RandomForest/mtbsiModel_RF.rds"))
saveRDS(mtbsiModel_RF, file.path(analyzed_src, "RandomForest/mtbsiModel_RF_2.rds"))
saveRDS(mtbsiModel_RF, file.path(analyzed_src, "RandomForest/mtbsiModel_RF_3.rds"))

# ================================================================
# Load save model
mtbsiModel_RF <- readRDS(file.path(here("external/analyzed_data/RandomForest/mtbsiModel_RF_2.rds")))


# Print model
print(mtbsiModel_RF)


# Make predictions on the test dataset
pred_model_mtbsi <- predict(mtbsiModel_RF, newdata = test_data_mtbsi[,-1])

# Calculate the confusion matrix
conf_matrix_mtbsi <- confusionMatrix(pred_model_mtbsi, as.factor(test_data_mtbsi$severity))


# Plot the Random Forest model performance
plot(mtbsiModel_RF, main = "Model performance of RandomForest model for fire severity")


# Plot the importance of variables
important_variables_mtbsi <- varImp(mtbsiModel_RF, scale = T)
plot(important_variables_mtbsi, 
     type = "p", 
     main = "Feature importance plot using RF model for predicting fire severity",
     ylab = "Features")
```

## 9.7 XgBoost Modelling

### 9.7.1 xgboost modeling for predicting fire

```{r XGBoost for fire prediction}
library(xgboost)
library(caret)
cl <- makeCluster(4)
registerDoParallel(cl)

# Set up training control
train_control <- trainControl(
    method = "repeatedcv",
    number = 10,
    repeats = 5,
    verboseIter = TRUE,
    allowParallel = TRUE,
    classProbs = TRUE,
)
  

# Create more focused parameter grid based on typical ranges
tuneGrid <- expand.grid(
    nrounds = 3000,  
    max_depth = c(3, 5, 7),  # Common effective depths
    eta = 0.01,  # Learning rate options
    gamma = c(0, 0.1, 0.3),  # Tree complexity penalty
    colsample_bytree = c(0.8, 0.9, 1.0),  # Feature sampling
    min_child_weight = c(1, 3, 5),  # Minimum sum of instance weight
    subsample = c(0.8, 0.9, 1.0)  # Row sampling
)


train_data_FNF <- as.data.frame(train_data_FNF)
train_data_FNF
# Train model with progress printing
set.seed(42)  # For reproducibility
XGBmodel <- train(
    x = train_data_FNF[, predictors3],
    y = train_data_FNF[[targetFNF]],
    method = "xgbTree",
    trControl = train_control,
    tuneGrid = tuneGrid,
    metric = "Accuracy",
    verbose = TRUE,
    nthread = 1)

# Clean up parallel cluster
stopCluster(cl)


# Print best parameters and performance
print(XGBmodel$bestTune)
print(max(XGBmodel$results$Accuracy))

# Save model for later use
saveRDS(XGBmodel, "XGBmodel_final.rds")
saveRDS(XGBmodel, "XGBmodel_final_3.rds")


# Load save model
XGBmodel <- readRDS(file.path(here("XGBmodel_final_3.rds")))
XGBmodel



# Make new predictions
xgbpred <- predict(XGBmodel, newdata = test_data_FNF)

predictions_df <- data.frame(
  predicted_values = xgbpred)
predictions_df

# Classification metrics
confusion_matrix <- confusionMatrix(xgbpred, test_data_FNF[[targetFNF]])
performance <- list(
  confusion_matrix = confusion_matrix,
  accuracy = confusion_matrix$overall["Accuracy"],
  kappa = confusion_matrix$overall["Kappa"]
)


# Get variable importance
importance <- varImp(XGBmodel)$importance
importance$Features <- rownames(importance)

importance

# Plot the importance of variables
ggplot(importance, aes(x = reorder(Features, Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Feature Importance Plot using XGBoost model for predicting Fire",
    x = "Features",
    y = "Importance %"
  )



# Prepare final results
results <- list(
  model = XGBmodel,
  performance = performance,
  importance = importance,
  predictions = predictions_df,
  actual = test_data_FNF[[targetFNF]],
  train_data = train_data_FNF,
  test_data = test_data_FNF,
  best_tune = XGBmodel$bestTune
)


results
```

### 9.7.2 xgboost modeling for predicting fire severity

```{r XGBoost for fire severity}
library(xgboost)
library(caret)

cl <- makeCluster(4)
registerDoParallel(cl)

# Set up training control
train_control <- trainControl(
    method = "repeatedcv",
    number = 10,
    repeats = 5,
    verboseIter = TRUE,
    allowParallel = TRUE,
    classProbs = TRUE,
)
  

# Create more focused parameter grid based on typical ranges
tuneGrid <- expand.grid(
    nrounds = 3000,  
    max_depth = 5,  # Common effective depths
    eta = 0.01,  # Learning rate options
    gamma =  0.3,  # Tree complexity penalty
    colsample_bytree = 0.8,  # Feature sampling
    min_child_weight = 1,  # Minimum sum of instance weight
    subsample = 1.0 # Row sampling
)


train_data_mtbsi <- as.data.frame(train_data_mtbsi)
train_data_mtbsi
# Train model with progress printing
set.seed(42)  # For reproducibility
XGBmodel <- train(
    x = train_data_FNF[, predictors],
    y = train_data_FNF[[targetFNF]],
    method = "xgbTree",
    trControl = train_control,
    tuneGrid = tuneGrid,
    metric = "Accuracy",
    verbose = TRUE,
    nthread = 1)

# Clean up parallel cluster
stopCluster(cl)


# Print best parameters and performance
print(XGBmodel$bestTune)
print(max(XGBmodel$results$Accuracy))

# Save model for later use
saveRDS(XGBmodel, file.path(analyzed_src, "xgboost/fireSeverity_Prediction.rds"))

XGBmodel

# Make new predictions
xgbpred <- predict(XGBmodel, newdata = test_data_FNF)

predictions_df <- data.frame(
  predicted_values = xgbpred)
predictions_df

# Classification metrics
confusion_matrix <- confusionMatrix(xgbpred, test_data_FNF[[targetFNF]])
performance <- list(
  confusion_matrix = confusion_matrix,
  accuracy = confusion_matrix$overall["Accuracy"],
  kappa = confusion_matrix$overall["Kappa"]
)


# Get variable importance
importance <- varImp(XGBmodel)$importance
importance$Features <- rownames(importance)

importance

# Plot the importance of variables
ggplot(importance, aes(x = reorder(Features, Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Feature Importance Plot using XGBoost model for predicting Fire",
    x = "Features",
    y = "Importance %"
  )



# Prepare final results
results <- list(
  model = XGBmodel,
  performance = performance,
  importance = importance,
  predictions = predictions_df,
  actual = test_data_FNF[[targetFNF]],
  train_data = train_data_FNF,
  test_data = test_data_FNF,
  best_tune = XGBmodel$bestTune
)



```

### 

## 9.8 GAM model

```{r GAM model}
library(mgcv)
cl <- makeCluster(4)
registerDoParallel(cl)


# Define the control for cross-validation
train_control <- trainControl(method = "repeatedcv",
                            number = 10,
                            repeats = 5,
                            allowParallel = T)

# Define tuning grid for GAM
# Adjust select and bs parameters based on your features
gam_grid <- expand.grid(
    select = c(T, F),
    method = "GCV.Cp",
    bs = "tp"  # thin plate regression splines
)

# Train the GAM model
set.seed(123)
gam_model <- train(plotLabel ~ .,
                  data = train_data_FNF,
                  method = "gam",
                  trControl = train_control,
                  tuneGrid = gam_grid,
                  family = binomial(link = "logit"))  # adjust family based on your response variable

stopCluster(cl)
  
```

```{r}
library(data.table)
outbreakDT <- as.data.table(outbreak_lodgepole)

```
