---
title: "Manuscript_v1.01"
author: "S. Paudel"
date: "2024-12-20"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




# A. Load packages
```{r}
# Install forDistR packages from github
if (!require("remotes")) {
  install.packages("remotes")
}

tryCatch({
  remotes::install_github(
    "paudelsushil/forestDistR",
    ref = "main",
    dependencies = TRUE,
    upgrade = "never"
  )
}, error = function(e) {
  message("Error installing package: ", e$message)
})


if (!require("drat")) {
  install.packages("drat")
 }
drat::addRepo("cloudyr", "http://cloudyr.github.io/drat")


# List of libraries
libs <- c("here", "ranger", "tuneRanger", "caret","tidyverse", "arrow", "terra", "sf", "sp", "zoo", "data.table", "ncdf4", "gridExtra", "ggspatial", "patchwork", "magick")


# List packages that are not installed
installedLibs <- libs %in% 
  rownames(
    installed.packages()
    ) 
# Install packages if not installed
if(any(installedLibs == F)){
  install.packages(
    libs[!installedLibs]
  )
}
                   
# Laod packages
base::invisible(
  lapply(
    libs, library, character.only = T
  )
)        


```


# B. Load working directories
```{r}
# Define Raw datasets location
raw_src <- here::here("external/raw_data")
# Define analyzed datasets location
analyzed_src <- here::here("external/analyzed_data")

# Define the directory for the raw data
tmmxM_dir <- here::here("external/analyzed_data/climate_data", "tmmx", "monthly")
tmmnM_dir <- here::here("external/analyzed_data/climate_data", "tmmn", "monthly")
prM_dir <- here::here("external/analyzed_data/climate_data", "pr", "monthly")

pdsiM_dir <- here::here("external/raw_data/climate_data", "pdsi")
scPdsiM_dir <- here::here("external/raw_data/climate_data", "scpdsi")
pzIM_dir <- here::here("external/raw_data/climate_data", "pzi")

vpdM_dir <- here("external/analyzed_data/climate_data/vpd/monthly")
petM_dir <- here("external/analyzed_data/climate_data/pet/monthly")
vsM_dir <- here("external/analyzed_data/climate_data/vs/monthly")
rmaxM_dir <- here("external/analyzed_data/climate_data/rmax/monthly")
rminM_dir <- here("external/analyzed_data/climate_data/rmin/monthly")
sradM_dir <- here("external/analyzed_data/climate_data/srad/monthly")

tpi_dir <- here("external/analyzed_data/topographic")
twi_dir <- here("external/analyzed_data/topographic")
elev30m_dir <- here("external/analyzed_data/topographic")
slpD30m_dir <- here("external/analyzed_data/topographic")
aspD30m_dir <- here("external/analyzed_data/topographic")

awC_dir <- here("external/analyzed_data/topographic/awC")
# Define the directory for the analyzed data

# nEW   version 
# 
filename <- "v25_5_15"
if(!dir.exists(file.path(analyzed_src, filename))){
  dir.create(file.path(analyzed_src, filename))
}

# results
if(!dir.exists(file.path(analyzed_src, filename, "results"))){
  dir.create(file.path(analyzed_src, filename, "results"))
}
# figures
if(!dir.exists(file.path(analyzed_src, filename, "results/figures"))){
  dir.create(file.path(analyzed_src, filename, "results/figures"))
}
# tables
if(!dir.exists(file.path(analyzed_src, filename, "results/tables"))){
  dir.create(file.path(analyzed_src, filename, "results/tables"))
}
# plots
if(!dir.exists(file.path(analyzed_src, filename, "results/plots"))){
  dir.create(file.path(analyzed_src, filename, "results/plots"))
}


# Output files source
results_src <- here::here(analyzed_src, filename, "results") 
fig_src <- here::here(results_src, "figures")
tbl_src <- here::here(results_src, "tables")
plots_src <- here::here(results_src, "plots")
```


# C. Load local functions
```{r}
source(here::here("R/functions.R"))
# source(here::here("R/getGridmet.metadata.R"))
# source(here::here("R/getGridmet.data.R"))
# source(here::here("R/gridmetDaily2Monthly.R"))
source(here::here("R/calculate.timeclass.R"))

source(here::here("R/extractValue2Points.R"))

source(here::here("R/get_cpu_info.R"))

source(here::here("R/get_ram_info.R"))

source(here::here("R/get_optimal_workers.R"))

source(here::here(r"(R\theme.R)"))

source(here::here(r"(R\corrPlot.R)"))

source(here::here("R/vif_plot.R"))
source(here::here("R/create_binary_feature.R"))
source(here::here("R/create_disturbance_features.R"))
source(here::here("R/analyze_vif_models.R"))
source(here::here("R/raster2table.R"))

```

# D. Administrative Extent
```{r}
conusStates <- sf::st_read(file.path(analyzed_src, 
                        "conus_states/conus_states.gpkg"),
                        layer = "conus_states")
```
# 1. Lodgepole Extent
```{r}
lodgepoleExtent <- sf::st_read(file.path(here("external/analyzed_data/study_area/lodgepoleExtent.gpkg"))) %>% 
                    sf::st_set_agr("identity")            


lodgepoleExtent <- lodgepoleExtent %>% 
                    mutate(studyArea = "lodgepoleExtent",
                           studyA_ha = calculate.area(., "ha")) %>% 
                    st_transform("EPSG:5070") %>% 
                    sf::st_set_agr("identity")                        

lodgepoleExtent
# sum(studyArea$studyA_ha)
```


# Elevation
```{r}
# elev30m <- terra::rast(file.path(here("external/analyzed_data/topographic/elev30m.tif"))) 



# # Convert to Cloud Optimized GeoTiff
# gdalUtilities::gdal_translate(
#     src_dataset = file.path(here("external/analyzed_data/topographic/Extract_LC201.tif")),
#     dst_dataset = file.path(here("external/analyzed_data/topographic/elev30m.tif")),
#     of = "GTiff",
#     co = c("TILED=YES", 
#            "COPY_SRC_OVERVIEWS=YES", 
#            "COMPRESS=DEFLATE",
#            "PREDICTOR=2",
#            "ZLEVEL=9"))



```

# 2. Fire Extent
```{r}
# Load fire perimeter
fireLodgepole <- sf::st_read(file.path("external/analyzed_data/mtbs_fire_data/fireLodgepole.gpkg")) 
glimpse(fireLodgepole)


# fire_disturbancePerYr <- fireLodgepole %>% group_by(fireYear) %>% summarise(totalFireA_ha = sum(FireA_ha))


# Filter and manupulate data
# fireLodgepole <- fireLodgepole %>% 
#                     rename_with(tolower) %>% 
#                     dplyr::filter(incid_type == "Wildfire") %>%
#                     mutate(fireDate = lubridate::as_date(ig_date),
#                             fireYear = lubridate::year(ig_date)) %>%
#                     rename(geom = shape) %>%  
#                     st_transform("EPSG:5070") %>% 
#                     dplyr::select(event_id, fireDate, fireYear, geom) %>% 
#                     mutate(FireA_ha = calculate.area(., area.unit = "ha")) %>% 
#                     mutate(FireA_km2 = calculate.area(., area.unit = "km2")) %>% 
#                     calculate.timeclass("fireYear", 1984, 2023, 10, "fire") %>% 
#                     sf::st_set_agr("identity")
fireLodgepole

# st_write(fireLodgepole, file.path("external/analyzed_data/mtbs_fire_data/fireLodgepole.gpkg"), append = FALSE)

```

# 3. Fire Severity
## 3.1 Masking and reprojecting mtbs severity maps
```{r}
# # Masking 
# rawMtbsI_src <- dir(file.path("external/raw_data/mtbs_fire_data/burnSeverity_tifs"), pattern = ".tif", full.names = T)
# output_src <- file.path(analyzed_src, "mtbs_fire_data/mtbs_severity")

# # Transform to crs of raster data 
# vecConusStates <- st_transform(conusStates, st_crs(rast(rawMtbsI_src[40])))

# # clip burn severity
# raster_clip(rawMtbsI_src[40], vecConusStates, output_src)

# # Western CONUS burn severity raster data
# conusMtbsI_src <- dir(file.path(analyzed_src, "mtbs_fire_data/mtbs_severity"), pattern = ".tif", full.names = T)


# # Reproject fire severity raster for 2023
# mtbs_CONUS_2023 <- project(rast(conusMtbsI_src[40]), "EPSG:5070")

# # Write a raster to the file
# filename <- file.path(analyzed_src, "mtbs_fire_data/mtbs_severity", "conus_epsg5070", "mtbs_CONUS_2023.tif")
# filename
# writeRaster(mtbs_CONUS_2023, filename, overwrite=TRUE)

# conusMtbsI_src


```

## 3.2 Extract severity for each fire perimeter using centroids

```{r}




# mtbs_2023 <- terra::rast(file.path(rastDir, "mtbs_CONUS_2023.tif")) %>% 
# terra::project("epsg:5070")

# # Save the reprojected raster
# writeRaster(mtbs_2023, 
#             file.path(rastDir, "mtbs_CONUS_2023_proj.tif"),
#             overwrite=TRUE)





# write a file
# st_write(fireSeverity, file.path(here::here("external/analyzed_data", filename, "fireSeveritySamples.gpkg")), append = F)



```



## 3.3 Fire Severity Plot (figure8)
```{r}
# plot_a <- plot.fireSeverity(fireSeverityInfo, periodic = F)
# plot_b <- plot.fireSeverity(fireSeverityInfo, periodic = T)


# # Combine the plots
# combined_plot <- grid.arrange(plot_a, plot_b,
#                               ncol = 1,
#                               heights = c(1, 1))
# # Save the plot
# ggsave(file.path(fig_src, "figure8.png"), combined_plot,
#        width = 7, height = 6, dpi = 300)
```

# 4. Outbreak (MPB)
```{r}
outbreakLodgepole <- sf::st_read(file.path("external/analyzed_data/outbreaks_data/outbreakLodgepole.gpkg"))

# mpbOutbreak <- sf::st_read(file.path("external/analyzed_data/outbreaks_data/mpbOutbreak_conus.gpkg"))
# mpbOutbreak <- mpbOutbreak %>% 
#         mutate(OutbreakA_ha = calculate.area(., area.unit = "ha"))
            




# plot(st_geometry(outbreakLodgepole))

outbreakLodgepole <- outbreakLodgepole %>% 
                        rename_with(tolower) %>% 
                        dplyr::filter(outbreak_agents == "mpb") %>%
                        rename(geom = shape, outbreakYear = outbreak_year) %>%  
                        dplyr::select(region_id, outbreakYear, state, geom) %>% 
                        mutate(OutbreakA_ha = calculate.area(., area.unit = "ha")) %>% 
                        calculate.timeclass("outbreakYear", 1984, 2023, 10, "mpb") %>% 
                        sf::st_set_agr("identity")

st_write(outbreakLodgepole, file.path("external/analyzed_data/outbreaks_data/outbreakLodgepole.gpkg"), append = FALSE)
table(outbreakLodgepole$mpbBy10yrs)


```

# 5. Fuel Management
```{r}
# Load fuel management dataset
fuelLodgepole <- sf::st_read(file.path("external/analyzed_data/fuel_mgmt/fuelLodgepole.gpkg"))
glimpse(fuelLodgepole)



# Fuel data manupulation
fuelLodgepole <- fuelLodgepole %>% 
                 mutate(fuelDate = lubridate::as_date(date_completed),
                        fuelYear = lubridate::year(date_completed)) %>%
                 rename(treatments = JFSP_CODE, stateABBR = state_abbr) %>%        
                 dplyr::select(stateABBR, fuelDate, fuelYear, treatments, activity, geom) %>% 
                 mutate(FuelA_ha = calculate.area(., area.unit = "ha"),
                        fuelMgmt = "Managed") %>% 
                 calculate.timeclass("fuelYear", 1984, 2023, 10, "fuel")  %>% 
                 sf::st_set_agr("identity")

st_write(fuelLodgepole, file.path("external/analyzed_data/fuel_mgmt/fuelLodgepole.gpkg"), append = FALSE)
```

# 6. Fire strata
## 6.1  Prepare Fire Strata

```{r}
# Fire and Outbreak
# fireOutbreak <- sf::st_intersection(fireLodgepole, outbreakLodgepole) %>% 
#                 st_make_valid(.) %>% 
#                 mutate(FireA_ha = calculate.area(., area.unit = "ha")) %>% 
#                 sf::st_cast("MULTIPOLYGON") %>%
#                 sf::st_set_agr("identity")

# # Write a data to file
# st_write(fireOutbreak, file.path(r"(external\analyzed_data\fireStrata)", "fireOutbreak.gpkg"), append = FALSE)

# # Fire and Fuel
# fireFuel <- sf::st_intersection(fireLodgepole, fuelLodgepole) %>% 
#             sf::st_make_valid(.) %>% 
#             mutate(FireA_ha = calculate.area(., area.unit = "ha")) %>% 
#             sf::st_cast("MULTIPOLYGON") %>% 
#             sf::st_set_agr("identity")

# # Write a data to file
# st_write(fireFuel, file.path("external/analyzed_data/fireStrata", "fireFuel.gpkg"), append = FALSE)

# # Fire, Fuel and Outbreak
# fireOutbreakFuel <- sf::st_intersection(fireOutbreak, fuelLodgepole) %>% 
#                     sf::st_make_valid(.) %>% 
#                     mutate(FireA_ha = calculate.area(., area.unit = "ha")) %>% 
#                     sf::st_cast("MULTIPOLYGON") %>%
#                     sf::st_set_agr("identity")

# # Write a data to file
# st_write(fireOutbreakFuel, file.path(r"(external\analyzed_data\fireStrata)", "fireOutbreakFuel.gpkg"), append = FALSE)


# # Plot the frequencies of fire extents in each stratum
# par(mfrow = c(3, 1))
# hist(fireOutbreak$FireA_ha)
# hist(fireFuel$FireA_ha)
# hist(fireOutbreakFuel$FireA_ha)



table(fireLodgepole$fireYear)
```

# figure 1
```{r}
# Panel A: MPB outbreaks over time
plot_a <- ggplot(outbreakLodgepole, aes(x = factor(outbreakYear), y = OutbreakA_ha)) +
  geom_bar(stat = "identity", fill = "#1f77b4", width = 0.8) +
  labs(x = NULL,  # remove x label as it's shared
       y = expression("Area (ha × 10"^"3"*")"),
       title = "(a) Mountain pine beetle (MPB) outbreaks") +
  plot_theme() +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = function(x) x/1000,  # Convert to thousands
                       expand = c(0, 0))
plot_a  
# Panel B: Fires over time
plot_b <- ggplot(fireLodgepole, aes(x = factor(fireYear), y = FireA_ha)) +
  geom_bar(stat = "identity", fill = "#ff7f0e", width = 0.8) +
  labs(x = NULL,  # remove x label as it's shared
       y = expression("Area (ha × 10"^"3"*")"),
       title = "(b) Fire") +
  plot_theme() +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = function(x) x/1000,  # Convert to thousands
                       expand = c(0, 0))
plot_b
# Panel C: Fuel management over time
plot_c <- ggplot(fuelLodgepole, aes(x = factor(fuelYear), y = FuelA_ha)) +
  geom_bar(stat = "identity", fill = "#2ca02c", width = 0.8) +
  labs(x = "Year",
       y = expression("Area (ha × 10"^"3"*")"),
       title = "(c) Fuel management") +
  plot_theme() +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = function(x) x/1000,  # Convert to thousands
                       expand = c(0, 0))

# Panel D: Annual mean, maximum, and Minimum PDSI values
pdsiM_df <- raster2table_annual(pdsiM_dir, "\\.nc$", "pdsiM", aoi = conusStates)

# vpdM_df <- raster2table_annual(vpdM_dir, "\\.nc$", "vpdM", aoi = conusStates)


plot_d <- plot_climate_trends(pdsiM_df, "PDSI")

plot_d
# Combine the plots
combined_plot <- grid.arrange(plot_a, plot_b, plot_c, plot_d,
                            ncol = 2, 
                            heights = c(1, 1))
combined_plot

# Save the plot
ggsave(file.path(fig_src, "figure2.png"), combined_plot, 
       width = 7, height = 5, dpi = 500)
```

# figure 2 Study area maps
```{r}

maps_src <- file.path(results_src, 'maps')

plot_a <- magick::image_read(file.path(maps_src, 'lodgepole_map.jpg'))
plot_b <- magick::image_read(file.path(maps_src, 'mpb_map.jpg'))
plot_c <- magick::image_read(file.path(maps_src, 'fire_map.jpg'))
plot_d <- magick::image_read(file.path(maps_src, 'fuel_map.jpg'))

combined_map <- magick::image_append(c(plot_b, plot_c, plot_d ))

magick::image_write(combined_map, file.path(maps_src, "figure2.jpg"), quality = 100)

```


# figure 3
## figure 3(a) Outbreaks in managed Vs. un-managed

```{r, eval = T, echo = FALSE}
# Total outbreak over the time
totalOutbreaks <- sf::st_read(file.path(analyzed_src, "fireStrata/totalOutbreak.gpkg"))

totalOutbreak <- totalOutbreaks %>% 
  mutate(totalOutbreakA_ha = calculate.area(., "ha")) %>% 
  st_drop_geometry(.) %>% 
  dplyr::select(outbreak_year, outbreak_agents, totalOutbreakA_ha)


totalOutbreak

# Total outbreak on managed forest over the time
outbreakFuel <- sf::st_read(file.path(analyzed_src, "fireStrata/outbreakFuel.gpkg"))


# Disturbances in Managed forests
managedOutbreak <- outbreakFuel %>% 
  mutate(fuelMgmt = "Managed",
         managedOutbreakA_ha = calculate.area(., "ha")) %>% 
  dplyr::select(outbreak_year, outbreak_agents, fuelMgmt,  managedOutbreakA_ha, geom) %>% 
  st_drop_geometry(.)

managedOutbreak


# Summarize outbreak affected areas
outbreakFuelInfo <- full_join(
  totalOutbreak %>% 
    select(outbreak_year, totalOutbreakA_ha),
  managedOutbreak %>% 
    select(outbreak_year, managedOutbreakA_ha),
  by = "outbreak_year"
) %>%
  # Create long format with managed and unmanaged areas
  mutate(unmanagedOutbreakA_ha = totalOutbreakA_ha - managedOutbreakA_ha) %>%
  pivot_longer(
    cols = c(managedOutbreakA_ha, unmanagedOutbreakA_ha),
    names_to = "mgmtStatus",
    values_to = "OutbreakA_ha"
  ) %>%
  mutate(
    mgmtStatus = case_when(
      mgmtStatus == "managedOutbreakA_ha" ~ "Managed",
      mgmtStatus == "unmanagedOutbreakA_ha" ~ "Unmanaged"
    )
  )
outbreakFuelInfo
outbreakFuelInfo <-  outbreakFuelInfo %>% 
  mutate(P_OutbreakA_ha = OutbreakA_ha/sum(OutbreakA_ha))


sum(outbreakFuelInfo$P_OutbreakA_ha)
# sum(outbreakLodgepole$OutbreakA_ha)

```

## figure 3(b) Fire in managed vs. un-managed

```{r Fire in Managed vs Unmanaged, eval = T, echo = F}

# Total outbreak over the time
totalFire <- sf::st_read(file.path(analyzed_src, "fireStrata/totalFire.gpkg"))
totalFire
table(totalFire$fireYear)

totalFireInfo <- totalFire %>% 
  mutate(totalFireA_ha = calculate.area(., "ha"))  %>% 
  st_drop_geometry(.) %>% 
  dplyr::select(fireYear, totalFireA_ha)


# totalFireA_ha <- sum(totalFireInfo$totalFireA_ha)
totalFireA_ha
totalFireInfo
# Total outbreak on managed forest over the time
fireFuel <- sf::st_read(file.path(analyzed_src, "fireStrata/fireFuel.gpkg"))
fireFuel

# Disturbances in Managed forests
managedFire <- fireFuel %>% 
  mutate(fuelMgmt = "Managed",
         managedFireA_ha = calculate.area(., "ha")) %>% 
  dplyr::select(fireYear, fuelMgmt,  managedFireA_ha) %>% 
  group_by(fireYear) %>%
  summarise(managedFireA_ha = sum(managedFireA_ha)) %>%
  st_drop_geometry(.) %>% 
  rbind(., data.frame(fireYear = c(1995, 1997), managedFireA_ha = c(0, 0)))

managedFire

# Summarize outbreak affected areas
fireFuelInfo <- full_join(
  totalFireInfo %>% 
    dplyr::select(fireYear, totalFireA_ha),
  managedFire %>% 
    dplyr::select(fireYear, managedFireA_ha),
  by = "fireYear"
) %>%
  # Create long format with managed and unmanaged areas
  mutate(unmanagedFireA_ha = totalFireA_ha - managedFireA_ha) %>%
  pivot_longer(
    cols = c(managedFireA_ha, unmanagedFireA_ha),
    names_to = "mgmtStatus",
    values_to = "FireA_ha"
  ) %>%
  mutate(
    mgmtStatus = case_when(
      mgmtStatus == "managedFireA_ha" ~ "Managed",
      mgmtStatus == "unmanagedFireA_ha" ~ "Unmanaged"
    )
  )

# table(is.na(fireFuelInfo$FireA_ha))

# fireFuelInfo <- fireFuelInfo %>% mutate(P_FireA_ha = FireA_ha/sum(FireA_ha))
fireFuelInfo
sum(fireFuelInfo$FireA_ha)
```

## figure(3c) Fire in managed: with vs without outbreaks

```{r}
# Fire AND Fuel AND Outbrea, 
fireOutbreakFuel <- sf::st_read(file.path(analyzed_src, "fireStrata/fireOutbreakFuel.gpkg"))
totalMPB_dff_fireOutbreakFuel <- sf::st_read(file.path(analyzed_src, "fireStrata/totalOutbreak_DIFF_fireOutbreakFuel.gpkg"))

colnames(fires_no_outbreak)

str(fireOutbreakFuel)

  
fireOutbreakFuel <- fireOutbreakFuel %>%
  mutate(
    outbreak_status = "Outbreak",
    fire_outbreak_area_ha = calculate.area(., "ha")
  )  %>% 
  dplyr::select(outbreakYear, outbreak_status, fire_outbreak_area_ha) %>%
  st_drop_geometry(.)

# Get fires without outbreaks using st_difference
fires_no_outbreak <- totalMPB_dff_fireOutbreakFuel %>%
  mutate(
    outbreak_status = "No Outbreak",
    fire_outbreak_area_ha = calculate.area(., "ha")
  ) %>%
  dplyr::select(outbreak_year, outbreak_status, fire_outbreak_area_ha) %>%
  st_drop_geometry(.) %>% 
  rename(outbreakYear = outbreak_year)

# Combine the data
fire_outbreak_summary <- bind_rows(fireOutbreakFuel, fires_no_outbreak) %>%
  group_by(outbreakYear, outbreak_status) %>%  # Added fire_year to grouping
  summarize(
    total_area_ha = sum(fire_outbreak_area_ha),
    .groups = "drop"
  ) %>%
  group_by(outbreakYear) %>%  # Group by year to calculate percentages within each year
  mutate(
    percent_area = total_area_ha / sum(total_area_ha)
  ) %>%
  ungroup()

fire_outbreak_summary

 
```

# Figure 3(a, b, & c)

```{r figure3}
# Panel A: Percent area affected by Outbreaks in managed vs unmanaged forests by year
 
plot_3a <- ggplot(outbreakFuelInfo, 
                 aes(x = factor(mgmtStatus), y = OutbreakA_ha/sum(OutbreakA_ha), fill = mgmtStatus)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  labs(x = "",
       y = expression("% of forests affected by outbreaks"),
       fill = "Forests",
       title = "(a) % of forests affected by outbreaks in managed forests vs. unmanaged forests") +
    
  plot_theme() +
  scale_fill_brewer(palette = "Set2") +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = scales::percent_format())

plot_3a

# Panel B: Percent area affected by Fires in managed vs unmanaged forests by year
plot_3b <- ggplot(fireFuelInfo, 
                 aes(x = factor(mgmtStatus), y = FireA_ha/sum(FireA_ha), fill = mgmtStatus)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  labs(x = "",
       y = expression("% of forests that burned"),
       fill = "Forests",
       title = "(b) % of forests affected by fires in managed forests vs. unmanaged forests") +
  plot_theme() +
  scale_fill_brewer(palette = "Set2") +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = scales::percent_format())

plot_3b
# Panel C: Fires in managed areas: with vs without subsequent outbreaks by year
fire_outbreak_summary
plot_3c <- fire_outbreak_summary %>%
  ggplot(aes( x = factor(outbreak_status), y = total_area_ha/sum(total_area_ha), fill = outbreak_status)) +
  # Bar plot configuration
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  labs(
    x = "",
    y = expression("% of forests that burned"),
    fill = "Outbreak Status",
    title = "(c) Fires in managed forests: with outbreaks vs. without outbreaks"
  ) +
  # Color scheme
  scale_fill_manual(
    values = c("No Outbreak" = "#2ca02c", "Outbreak" = "#d62728")) +
  # Labels and title
  plot_theme() +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = scales::percent_format())

plot_3c

# Combine the plots
combined_plot3 <- grid.arrange(plot_3a, plot_3b, plot_3c, 
                            ncol = 1, 
                            heights = c(1.5, 1.5, 1.5))
combined_plot3
# Save the plot
ggsave(file.path(fig_src, "figure3.png"), combined_plot3, 
       width = 7, height = 10, dpi = 300)

```


# Figure 4

```{r figure4}
fireFuelInfo
# Panel A: Percent area affected by Outbreaks in managed vs unmanaged forests
plot_4a <- ggplot(outbreakFuelInfo, aes(x = factor(outbreak_year), y = OutbreakA_ha/totalOutbreakA_ha, 
                                      fill = mgmtStatus)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  labs(x = "Year",
       y = expression("% of total study area"),
       fill = "Management Status",
       title = "(a) % of total forests affected by outbreaks in managed forests vs. unmanaged forests") +
  plot_theme() +
  scale_fill_brewer(palette = "Set2") +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = scales::percent_format())

plot_4a

# Panel B: Percent area affected by Fires in managed vs unmanaged forests
plot_4b <- ggplot(fireFuelInfo, aes(x = factor(fireYear), y = FireA_ha/totalFireA_ha, 
                                  fill = mgmtStatus)) +
  geom_bar(stat = "identity", position = "stack",  width = 0.8) +
  labs(x = "Year",
       y = expression("% of total study area"),
       fill = "Management Status",
       title = "(b) % of total forests affected by fires in managed forests vs. unmanaged forests") +
  plot_theme() +
  scale_fill_brewer(palette = "Set2") +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = scales::percent_format())
plot_4b   

# Panel C: Fires in managed areas: with vs without subsequent outbreaks
plot_4c <- ggplot(fire_outbreak_summary,                
                  aes(x = factor(outbreakYear), y = percent_area, fill = outbreak_status)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  scale_fill_manual(values = c("No Outbreak" = "#2ca02c", 
                              "Outbreak" = "#d62728"), 
                    labels = c("No Outbreak", "Outbreak")) +
  labs(x = "Year",
       y = expression("% of total of total study area"), 
       fill = "Outbreak Status",
       title = "(c) Fire in managed forests: with outbreaks vs. without outbreaks") +
  plot_theme() +
  scale_x_discrete(breaks = seq(1984, 2023, by = 5)) +
  scale_y_continuous(labels = scales::percent_format())
plot_4c


# Combine the plots
combined_plot4 <- grid.arrange(plot_4a, plot_4b, plot_4c, 
                            ncol = 1, 
                            heights = c(1, 1, 1))
combined_plot4
# Save the plot
ggsave(file.path(fig_src, "figure4.png"), combined_plot4, 
       width = 7, height = 8, dpi = 300)
```

## 6.2 Create fire presence samples for each stratum
```{r}
# Set to 1GB
options(future.globals.maxSize = 1000 * 1024^2) 

fireLodgepole
# Create samples of fire strata
set.seed(345)
fireSamples <- create.samples(fireLodgepole, "FireA_km2", type = "regular", n_cores = 10L, size = NULL)

fireSamples <- fireSamples %>% 
                mutate(event = "fire",
                       fireYear = lubridate::year(fireDate)) %>% 
                       rename(geom = x)
nrow(fireSamples)
table(fireSamples$fireYear)
nrow(fireSamples) - 17550
rastDir <-here::here("external/analyzed_data/mtbs_fire_data/mtbs_severity/conus_epsg5070")
# Exctract Fire Severity Index from the Raster data
fireSamples <- extractValue2Points(fireSamples,
                                        rastDir,
                                        variable = "mtbsI",
                                        match.column = "fireDate")

fireSamples <- fireSamples %>% 
  mutate(severity = case_when(
    mtbsI == 1 ~ "Unburned/Low",                           
    mtbsI == 2 ~ "Low",        
    mtbsI == 3 ~ "Moderate",            
    mtbsI == 4 ~ "High",             
    mtbsI == 5 ~ "Outside Burn Area", 
    mtbsI == 6 ~ "Unmappable", 
    TRUE ~ "Increased Greeness"
  ))

# Save a fire samples to a file
st_write(fireSamples, file.path(plots_src, "samples_km2.gpkg"))

fireSamples <- sf::st_read(file.path(plots_src, "fireSamples.gpkg")) 

glimpse(fireSamples)

table(fireSamples$severity)
# Monte Carlo approach
# n_iterations <- 4
# fireSamples_set <- list()

# for(i in 1:n_iterations) {
#   fireSamples_set[[i]] <- fireSamples %>%
#     group_by(disturbance) %>%
#     slice_sample(n = 585) %>%
#     ungroup() %>% 
#     mutate(set_num = i )
# }

# # Write to geopackages
# for(i in 1:length(fireSamples_set)) {
#   st_write(fireSamples_set[[i]], 
#   file.path("external/analyzed_data/plots/strata", paste0("fireSamples_set", i, ".gpkg")), 
#   append = FALSE)
# }
library(automap)
in
# table(fireSamples_set[[1]]$outbreak)
# table(fireSamples_set[[1]]$fuelmgmt)
# table(fireSamples_set[[1]]$fireStrata)


# table(fireStrataSmpls$strata)
33*1.154


```



## 6.3 Load fire Samples set from directory
```{r}
# n_sets <- 4 
# for(i in 1:n_sets) {
#     assign(paste0("fireSamples_set", i), 
#            sf::st_read(file.path("external/analyzed_data/plots/strata", 
#                                 paste0("fireSamples_set", i, ".gpkg"))))
# }



```



## 6.4 Randomly select nonFire Samples for each fire observations
```{r}
# Create allFireA
allFiresExtent <- st_union(fireLodgepole) %>% st_make_valid(.)
                 
lodgepoleExtent <- st_union(lodgepoleExtent) %>% st_make_valid(.)


# Create the non-Fire area within Lodgepole pine forests area.
nonFiresExtent <- st_difference(lodgepoleExtent, allFiresExtent) %>% st_make_valid(.)




# Fire date count to make a pair with nonFire points
fire_date_counts <- fireSamples %>%
  group_by(fireDate, severity) %>%
  summarise(n = n()) %>%
  arrange(fireDate)


# Create point for nonFireArea
set.seed(123)
nonFireSamples <- sf::st_sample(nonFiresExtent, size = nrow(fireSamples) * 2, type = "regular") %>% 
                                st_as_sf(.) %>% 
                                dplyr::filter(!rowSums(st_intersects(., allFiresExtent, sparse = F)) > 0) %>% 
                                slice_sample(n = nrow(fireSamples))

# Create non-fire samples with balanced distribution
set.seed(123)
nonFireSamples <- nonFireSamples %>% 
                   mutate(
                          fireDate = sample(
                          rep(fire_date_counts$fireDate, fire_date_counts$n),
                          size = n(),
                          replace = FALSE
                        )) %>%
                  mutate(
                        event = "non_fire",
                        fireYear = lubridate::year(fireDate)
                      ) %>% 
                      rename(geom = x)

nonFireSamples

samples <- dplyr::bind_rows(fireSamples, nonFireSamples)

samples


# Write a data to file
sf::st_write(samples, file.path(plots_src, "samples_km2.gpkg"))





glimpse(samples)

```

# 7. Sample Plots
```{r}
# Extract all the features value using samplePlots
samples <- st_join(samples, outbreakLodgepole, join = st_intersects, left = TRUE)
nrow(samples)

samples <- st_join(samples, fuelLodgepole, join = st_intersects, left = TRUE)
nrow(samples)

glimpse(samples)



# Only outbreak and fuel mgmt those are followed by fire are
# taken as present and rest are labelled as none disturbace.
samplePlots <- samplePlots %>% 
                mutate(X = st_coordinates(.)[, 1],
                       Y = st_coordinates(.)[, 2]) %>%
                mutate(MPB_lag = outbreakYear - fireYear,
                       fuel_lag = fuelYear - fireYear) %>%
                       
                mutate(outbreak = if_else(MPB_lag >= 0, 'mpb', 'no_mpb'),
                      fuel_mgmt = if_else(fuel_lag >= 0, 'fuel', 'no_fuel')) %>% 
                mutate(outbreak = if_else(is.na(outbreak), "no_mpb", "mpb"),
                        fuel_mgmt = if_else(is.na(fuel_mgmt), 'no_fuel', 'fuel')) %>% 
                mutate(mpb_lag_class = case_when(
                        MPB_lag >= 0 & MPB_lag <= 3  ~ "mpb_red",
                        MPB_lag >= 4 & MPB_lag <= 6 ~ "mpb_young_gray",
                        MPB_lag >= 7 & MPB_lag <= 14 ~ "mpb_old_gray",
                        MPB_lag > 14 ~ "mpb_old",
                        MPB_lag < 0 ~ "no_mpb"
                      )) %>%
                mutate(fuel_lag_class = case_when(
                        fuel_lag >= 0 & fuel_lag <= 5 ~ "fuel_0_5",
                        fuel_lag >= 6 & fuel_lag <= 10 ~ "fuel_6_10",
                        fuel_lag >= 11 & fuel_lag <= 15 ~ "fuel_11_15",
                        fuel_lag >= 16 ~ "fuel_gt_15",
                        fuel_lag < 0 ~ "no_fuel"
                      )) %>%
                mutate(severity = case_when(
                        severity == "Low" ~ "Low",
                        severity == "Moderate" ~ "Moderate",
                        severity == "High" ~ "High",
                        TRUE ~ "non_fire"
                      )) %>%                  
                mutate(fid = 1:n()) %>% 
                st_make_valid(.)      
samplePlots %>% filter(outbreak == "mpb" & mpb_old_gray == 1)
  

source(here::here("R/create_binary_feature.R"))
samplePlots <- create_binary_features(samplePlots, "mpb_lag_class", c("mpb_red", 'mpb_old_gray', "mpb_old", "mpb_young_gray"))
samplePlots <- create_binary_features(samplePlots, "fuel_lag_class", c("fuel_0_5",'fuel_6_10', 'fuel_11_15', 'fuel_gt_15'))

                               


# Write a data to file
st_write(samplePlots, file.path(plots_src, "samplePlots_v2.gpkg"), layer_options = c("FID=fid"), append = F)


colnames(samplePlots)


```



# 7.1 Splitting sample plots into 4 decades
```{r}
samplePlots <- sf::st_read(file.path("G:/ForDis/external/analyzed_data/v25_3_15/results/plots/samplePlots_v2.gpkg"))
# ss <- sf::st_read(file.path(plots_src, "samplePlots.gpkg"))                              
glimpse(samplePlots)


table(samplePlots$fireBy10yrs)
decades <- unique(samplePlots$fireBy10yrs)
decades
for(decade in decades) {
  # Extract start and end years from the decade label
  years <- unlist(strsplit(as.character(decade), "-"))
  start_year <- years[1]
  end_year <- years[2]
  
  # Subset the data for this decade
  decade_plots <- samplePlots[samplePlots$fireBy10yrs == decade,]
  
  # check dir exists
  if(!dir.exists(file.path(plots_src, "decadal_plots"))) {
    dir.create(file.path(plots_src, "decadal_plots"), recursive = TRUE)
  }
  # Create output filename
  output_file <- file.path(plots_src, "decadal_plots", 
                          sprintf("samplePlots_%s_%s.gpkg", start_year, end_year))
  
  # Write to file
  st_write(decade_plots, output_file, driver = "GPKG", append = FALSE)
}

# samplePlots <- samplePlots %>% sf::st_as_sf(.) %>% st_transform("epsg:5070")

fireSeverityPlots <- samplePlots %>% 
  filter(severity %in% c("High", "Moderate", "Low"))%>% 
  mutate(severity = factor(severity, levels = c("High", "Moderate", "Low")))


glimpse(fireSeverityPlots)

# Write a data to file
st_write(fireSeverityPlots, file.path(plots_src, "fireSeverityPlots.gpkg"), layer_options = c("FID=fid"), append = F)

```

# 8 Data Pre-Processing
## 8.1 
```{r}

# Target variable for fire and fire severity
response_var = c("event", "severity")

fCol <- c("fuel_mgmt", "outbreak")

facCols <- c("mpb_red", "mpb_young_gray", "mpb_old_gray",  "mpb_old",  
                "fuel_0_5",  "fuel_6_10",  "fuel_11_15",  "fuel_gt_15")    

numCol <- c("tmmxM", "tmmnM", "prM", "pdsiM",  "scPdsiM", "pzIM", 
          "vpdM", "petM", "vsM","rmaxM", "rminM", "sradM",
           "elev30M", "slpD30M", "aspD30M","tpi", "twi",
           "awC")

       
summary(samplePlots)               
colList <- c(facCols, numCol)
samplePlots <- fill
# Create a correlation matrix
# Coorelation plot
corrPlot(as.data.frame(samplePlots %>% st_drop_geometry(.)), numCol, 
        save_path = file.path(fig_src, "corrplot_v4.png"), 
         title = "Correlation plot of climate and topographic variables", 
         width = 6, height = 7)
#check numColl is in samplePLots
samplePlots %>% dplyr::select(all_of(numCol)) %>% str(.)



```


## 8.2 
```{r}

glimpse(samplePlots)         
# Dataset for fire nonfire
data_FNF <- process_fnf_data(samplePlots, c(response_var, facCols, numCol)) %>% 
  dplyr::select(-severity, -X, -Y)

table(samplePlots$mpb_red)
str(data_FNF)
# write.csv(data_FNF, file.path(tbl_src, "data_FNF_v4.csv"))

# Dataset for mtbsI
data_mtbsI <- process_decade_data(samplePlots, all_of(colList)) 
data_mtbsI <- create_binary_features(data_mtbsI, "severity", c("Low", "Moderate", "High")) %>% dplyr::select(-event)
str(data_mtbsI)


# write.csv(data_mtbsI, file.path(tbl_src, "data_mtbsI_v4.csv"))

```

## 8.3 
```{r}
# Split the data into training and testing sets
set.seed(246)
sample_FNF <- sample.int(n = nrow(data_FNF), size = floor(.70*nrow(data_FNF)), replace = F)
sample_mtbsi <- sample.int(n = nrow(data_mtbsI), size = floor(.70*nrow(data_mtbsI)), replace = F)

train_data_FNF <- data_FNF[sample_FNF, ]
test_data_FNF  <- data_FNF[-sample_FNF,]

train_data_mtbsi <- data_mtbsI[sample_mtbsi, ]
test_data_mtbsi  <- data_mtbsI[-sample_mtbsi, ]

str(train_data_mtbsi)

```

## 8.4 Features selection
```{r}
# # Prepare data
response_var_fnf = "event"
response_var_mtbsi = "severity"            
# predictors  <-  c("mpb_red", "mpb_young_gray", "mpb_old_gray",  "mpb_old",  
#                 "fuel_0_5",  "fuel_6_10",  "fuel_11_15",  "fuel_gt_15", 
#                 "pdsiM", "pzIM", "tmmxM", "prM", 
#                 "slpD30M", "aspD30M","tpi", "twi",
#                 "awC")
colList
predictors  <-  c("mpb_red", "mpb_young_gray", "mpb_old_gray",  "mpb_old",  
                "fuel_0_5",  "fuel_6_10",  "fuel_11_15",  "fuel_gt_15", 
                "pdsiM", "pzIM", 'tmmxM', 'prM',
                "slpD30M", "aspD30M","tpi", "twi",
                "awC")                

# "mpb_red", "mpb_young_gray", "mpb_old_gray",  "mpb_old",  
#                 "fuel_0_5",  "fuel_6_10",  "fuel_11_15",  "fuel_gt_15"
predictors_collection <- list(
 



  # "predictors_1" = c("mpb_red", "mpb_young_gray", "mpb_old_gray",  "mpb_old",  
  #                 "fuel_0_5",  "fuel_6_10",  "fuel_11_15",  "fuel_gt_15",
  #                 "prM", "scPdsiM", "pzIM",  "vpdM", 
  #                 "slpD30M", "aspD30M", "tpi", "twi",  "awC"),

  # # scPdsiM, pzIM, and tmmxM
  # "predictors_2" = c("mpb_red", "mpb_young_gray", "mpb_old_gray",  "mpb_old",  
  #                 "fuel_0_5",  "fuel_6_10",  "fuel_11_15",  "fuel_gt_15",
  #                 "prM", "scPdsiM", "pzIM", "tmmxM", 
  #                  "slpD30M", "aspD30M", "tpi", "twi", "awC"),

  # # pzIM, and tmmxM
  # "predictors_3" = c("mpb_red", "mpb_young_gray", "mpb_old_gray",  "mpb_old",  
  #                 "fuel_0_5",  "fuel_6_10",  "fuel_11_15",  "fuel_gt_15",
  #                  "scPdsiM", "pzIM", "vpdM", 
  #                 "slpD30M", "aspD30M", "tpi", "awC"),

  # # scPdsiM, and vpdM
  # "predictors_4" = c("fuel_mgmt", "outbreak",
  #                 "prM", "scPdsiM", "pzIM",  "vpdM", 
  #                 "slpD30M", "aspD30M", "tpi", "twi",  "awC"),

  # # pzIM, and vpdM
  # "predictors_5" =  c("fuel_mgmt", "outbreak",
  #                 "prM", "scPdsiM", "pzIM", "tmmxM", 
  #                  "slpD30M", "aspD30M", "tpi", "twi", "awC"),

  # # pzIM, and petM
  # "predictors_6" = c("fuel_mgmt", "outbreak",
  #                  "scPdsiM", "pzIM", "vpdM", 
  #                 "slpD30M", "aspD30M", "tpi", "awC"),

  # # scPdsiM, and petM
  # "predictors_7" = c("mpb_red", "mpb_young_gray", "mpb_old_gray",  "mpb_old",  
  #                 "fuel_0_5",  "fuel_6_10",  "fuel_11_15",  "fuel_gt_15",
  #                 "prM", "scPdsiM", "petM", "vsM", 
  #                 "elev30M", "slpD30M", "aspD30M", "tpi", "twi", "awC"),
  # # scPdsiM, and vpdM
  # "predictors_8" = c("mpb_red", "mpb_young_gray", "mpb_old_gray",  "mpb_old",  
  #                 "fuel_0_5",  "fuel_6_10",  "fuel_11_15",  "fuel_gt_15",
  #                 "scPdsiM", "vpdM", 
  #                 "slpD30M", "aspD30M", "tpi", "awC"),
  # # scPdsiM, and petM  
  # "predictors_9" = c("mpb_red", "mpb_young_gray", "mpb_old_gray",  "mpb_old",  
  #                 "fuel_0_5",  "fuel_6_10",  "fuel_11_15",  "fuel_gt_15",
  #                 "scPdsiM", "petM",  
  #                 "slpD30M", "aspD30M", "tpi", "awC"),
  # # scPdsiM, and prM
  # "predictors_10" = c("mpb_red", "mpb_young_gray", "mpb_old_gray",  "mpb_old",  
  #                 "fuel_0_5",  "fuel_6_10",  "fuel_11_15",  "fuel_gt_15",
  #                 "scPdsiM", "prM",  
  #                 "slpD30M", "aspD30M", "tpi", "awC") 
)    


```




## 8.5 Variable Inflation Factor (VIF) analysis
```{r}
# VIF analysis
predictors
source(here::here("R/logistic_regression.R"))
vif_fnf_analyze <- analyze_vif_models(train_data_FNF, 
                              response_var = "event",
                              predictors = predictors,
                              output_dir = file.path(fig_src, "vif_fnf"),
                              title = "VIF aanalysis for fire occurrence",
                              plot_dims = list(width = 7, height = 8))
vif_fnf_analyze
vif_severity_analyze <- analyze_vif_models(train_data_mtbsi, 
                              response_var = "High",
                              predictors = predictors,
                              output_dir = file.path(fig_src, "vif_severity_High_v4"),
                              title = "VIF analysis for fire severity",
                              plot_dims = list(width = 7, height = 8))
```

# 9. Modelling
## 9.1 Using RNGER package for RF(random forest)
```{r}
library(ranger)
library(tuneRanger)
library(mlr)
library(qs)
            
table(data_FNF$fuel_mgmt)
table(data_FNF$outbreak)
table(samplePlots$mpb_lag_class)

train_data_FNF_subset <- train_data_FNF %>%
    dplyr::select(event, all_of(predictors)) 


# Creating a task
fire_event_task <- mlr::makeClassifTask(
    id = "fire_event",
    data = as.data.frame(train_data_FNF_subset),  # Convert to data.frame
    target = "event"                        # Specify target variable
)


# # Hyperparameter tuning for fire occurrence modeling
tuneRanger(
    task = fire_event_task,
    measure = list(acc),
    iters = 100,
    iters.warmup = 50,
    time.budget = NULL,
    num.threads = 5,
    num.trees = 2000,
    parameters = list(
      replace = TRUE, 
      respect.unordered.factors = "partition"),
    tune.parameters = c("mtry", "min.node.size", "sample.fraction"),
    # save.file.path = "external/analyzed_data/RandomForest/ranger_tuning_results",
    build.final.model = TRUE,
    show.info = getOption("mlrMBO.show.info", TRUE)
)

# Random Forest model using Ranger package
rf_model_FNF <- ranger(
    formula = as.formula(paste("event", "~", paste(predictors, collapse = " + "))),
    data = train_data_FNF,
    num.threads = 5,
    mtry = 2,
    num.trees = 2000,
    importance = 'permutation',
    scale.permutation.importance = TRUE,
    local.importance = TRUE,
    write.forest = TRUE,
    probability = TRUE,
    min.node.size = 2,
    replace = TRUE,
    splitrule = "gini",
    sample.fraction = 0.8154535,
    seed = 12,
    verbose = TRUE,
    oob.error = TRUE,
    classification = TRUE,
    keep.inbag = TRUE
)

print(paste("OOB Error Rate:", round(rf_model_FNF$prediction.error, 4)))
print(rf_model_FNF)

#

# Learning Curves
error_rates <- numeric(10)
tree_counts <- seq(500, 2000, by = 100)

for (i in seq_along(tree_counts)) {
  temp_model <- ranger(
    formula = as.formula(paste("event", "~", paste(predictors, collapse = " + "))),
    data = train_data_FNF,
    num.threads = 4,
    num.trees = tree_counts[i],
    min.node.size = 2,
    sample.fraction = 0.8154535,
    mtry = 2,
    oob.error = TRUE,
    keep.inbag = TRUE,
    classification = TRUE,
    splitrule = "gini",
    importance = 'permutation',
    scale.permutation.importance = TRUE,
    local.importance = TRUE,
    replace = TRUE,
    seed = 12
  )
  error_rates[i] <- temp_model$prediction.error
}


fnf_learnerPlot <- plot(tree_counts, error_rates, type = "l",
     xlab = "Number of Trees", ylab = "OOB Error Rate",
     main = "Random Forest Model Learning Curve for predicting fire occurrence")

ggsave(file.path(fig_src, "FNF_learning_curve.png"), fnf_learnerPlot,
       dpi = 500, unit = "in", width = 6, height = 4)

table(train_data_FNF$mpb_old_gray)
table(test_data_FNF$mpb_old_gray)

# Save the model
qs::qsave(rf_model_FNF, file = file.path(analyzed_src, filename, "models/rf_fnf_models_v2", "FNF_pred1.qs"))
qs::qsave(rf_model_FNF, file = file.path(analyzed_src, filename, "models/rf_fnf_models_v2", "FNF_pred2.qs"))
qs::qsave(rf_model_FNF, file = file.path(analyzed_src, filename, "models/rf_fnf_models_v2", "FNF_pred3.qs"))
qs::qsave(rf_model_FNF, file = file.path(analyzed_src, filename, "models/rf_fnf_models_v2", "FNF_pred4.qs"))
qs::qsave(rf_model_FNF, file = file.path(analyzed_src, filename, "models/rf_fnf_models_v2", "FNF_pred5.qs"))
qs::qsave(rf_model_FNF, file = file.path(analyzed_src, filename, "models/rf_fnf_models_v2", "FNF_pred6.qs"))

# Load the model
FNF_pred1 <- qs::qread(file.path(analyzed_src, filename, "models/rf_fnf_models_v2", "FNF_pred1.qs"))
FNF_pred2 <- qs::qread(file.path(analyzed_src, filename, "models/rf_fnf_models_v2", "FNF_pred2.qs"))
FNF_pred3 <- qs::qread(file.path(analyzed_src, filename, "models/rf_fnf_models_v2", "FNF_pred3.qs"))
FNF_pred4 <- qs::qread(file.path(analyzed_src, filename, "models/rf_fnf_models_v2", "FNF_pred4.qs"))
FNF_pred5 <- qs::qread(file.path(analyzed_src, filename, "models/rf_fnf_models_v2", "FNF_pred5.qs"))
FNF_pred6 <- qs::qread(file.path(analyzed_src, filename, "models/rf_fnf_models_v2", "FNF_pred6.qs"))



source(here::here("R/evaluate_ranger_model.R"))
# Model Evaluation
rf_model_FNF_evaluation <- evaluate_ranger_model(
  model = rf_model_FNF,
  test_data = test_data_FNF,
  actual_values = test_data_FNF$event,
  output_dir = file.path(analyzed_src, filename, "model_evaluation"),
  model_name = "rf_model_FNF"
)
glimpse(data_FNF)


# Print model evaluation metrics
rf_model_FNF_evaluation$metrics
# Print confusion matrix
rf_model_FNF_evaluation$confusion_matrix


fnfVip <- vip::vip(rf_model_FNF,
                  
                   scale = TRUE, 
                   num_features = 25,
                   geom = "point", 
                   all_permutation = TRUE, 
                   aesthetics = list(color = "blue", size = 2.5)) +  
    geom_text(aes(label = sprintf("%.1f", Importance)), 
              hjust = -0.3,  # Adjust horizontal position
              size = 2,    # Text size
              color = "black") +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 9, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 7, color = "blue", hjust = 0.5),
        axis.text.y = element_text(size = 6),  # Smaller text for many variables
        axis.text.x = element_text(size = 6),
        axis.title = element_text(size = 8),
        # plot.margin = margin(0.5, 1, 0.5, 0.5, "pt") # c(top, right, bottom, left)
    ) +
    labs(title = "Variable importance plot for fire occurrence between 1984 and 2023",
        
        
        #  x = "Variables",
         y = "Mean Decrease Accuracy (MDA)")
    # scale_x_continuous(expand = c(0, 100))
fnfVip
ggsave(file.path(fig_src, "figure5.png"), fnfVip, 
       dpi = 500, unit = "in", width = 6, height = 8)


# Fire Severity
# ====================================================================================
# Creating a task
# fire_severity_task <- mlr::makeClassifTask(id = "fire_severity",
#                       data = as.data.frame(train_data_mtbsi),
#                       target = "event")

# # Hyperparameter tuning for fire severity modeling
# tuneRanger(
#     task = fire_severity_task,
#     measure = list(acc),
#     iters = 100,
#     iters.warmup = 50,
#     time.budget = NULL,
#     num.threads = 5,
#     num.trees = 5000,
#     parameters = list(
#       replace = TRUE, 
#       respect.unordered.factors = "partition"),
#     tune.parameters = c("mtry", "min.node.size", "sample.fraction"),
#     save.file.path = "external/analyzed_data/RandomForest/fireSeverity_tuning_results",
#     build.final.model = TRUE,
#     show.info = getOption("mlrMBO.show.info", TRUE)
# )

# train_data_mtbsi_encoded <- train_data_mtbsi %>%
#                             dplyr::select(severity, all_of(predictors_collection[[4]])) %>% 
#                             create_binary_features(., "severity", 
#                                                    c("Low", "Moderate", "High"))

# train_data_mtbsi_high <- train_data_mtbsi_encoded %>%
#                             dplyr::select(severity_High, all_of(predictors_collection[[4]])) %>% 
#                             mutate(severity_High = factor(severity_High, levels = c(0, 1)))

# train_data_mtbsi_mod <- train_data_mtbsi_encoded %>%
#                             dplyr::select(severity_Moderate, all_of(predictors_collection[[4]])) %>% 
#                             mutate(severity_Moderate = factor(severity_Moderate, levels = c(0, 1)))

# train_data_mtbsi_low <- train_data_mtbsi_encoded %>%
#                             dplyr::select(severity_Low, all_of(predictors_collection[[4]])) %>% 
#                             mutate(severity_Low = factor(severity_Low, levels = c(0, 1)))
# process_decade_data(
#   samplePlots
# )
predictors <- predictors_collection[[4]]
# Fire severity modeling
rf_model_mtbsi_high <- ranger(
    formula = as.formula(paste("High", "~", paste(predictors, collapse = " + "))),
    data = train_data_mtbsi,
    num.threads = 4,
    mtry = 2,
    num.trees = 2000,
    importance = 'permutation',
    scale.permutation.importance = TRUE,
    local.importance = TRUE,
    write.forest = TRUE,
    probability = TRUE,
    min.node.size = 2,
    replace = TRUE,
    splitrule = "gini",
    sample.fraction = 0.8154535,
    seed = 12,
    verbose = TRUE,
    oob.error = TRUE,
    classification = TRUE,
    keep.inbag = TRUE
)

# Save the model
qs::qsave(rf_model_mtbsi_high, file = file.path(analyzed_src, filename, "models/rf_mtbsi_models_v2", "mtbsi__high_pred1.qs"))
qs::qsave(rf_model_mtbsi_high, file = file.path(analyzed_src, filename, "models/rf_mtbsi_models_v2", "mtbsi_high_pred4.qs"))

# Read the model
mtbsi_pred4 <- qs::qread(file.path(analyzed_src, filename, "models/rf_mtbsi_models_v2", "mtbsi__high_pred4.qs"))

print(rf_model_mtbsi_high)

vip::vip(rf_model_mtbsi_high,
        scale = TRUE, 
        num_features = 20,
        geom = "point", 
        all_permutation = TRUE, 
        aesthetics = list(color = "blue", size = 2.5)) +  # Show top 20 variables
        geom_text(aes(label = sprintf("%.1f", Importance)), 
              hjust = -0.3,  # Adjust horizontal position
              size = 2,    # Text size
              color = "black") +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 10, face = "bold"),
        axis.text.y = element_text(size = 8),  # Smaller text for many variables
        axis.text.x = element_text(size = 8),
        axis.title = element_text(size = 10),
        plot.margin = margin(1, 1, 1, 1, "pt")
    ) +
    labs(title = "Variable Importance Plot for High wildfire severity between 1984 and 2023",
         x = "Variable",
         y = "Variables Importance Score")

# save vip plot
ggsave(file.path(fig_src, "figure7_vpd.png"), 
       dpi = 500, unit = "in", width = 6, height = 8)


# create vip plot for fire occurrence and fire severity
create_vip_table(
  models = list(rf_model_mtbsi_high, rf_model_mtbsi_high),
  model_name = c("Fire Occurrence", "Fire Severity"),
  output_dir = file.path(tbl_src, "tbls_vip_v4"),
  output_name = "table1_final"
  )

# Model Evaluation
rf_model_severity_evaluation <- evaluate_ranger_model(
  model = rf_model_mtbsi_high,
  test_data = test_data_mtbsi,
  actual_values = test_data_mtbsi$High,
  output_dir = file.path(analyzed_src, filename, "model_evaluation"),
  model_name = "rf_model_mtbsi_vpd"
)

samplePlots
rf_model_severity_evaluation$metrics

rf_model_mtbsi_mod <- ranger(
    formula = severity_Moderate ~ .,
    data = train_data_mtbsi_mod,
    num.threads = 4,
    mtry = 2,
    num.trees = 5000,
    importance = 'permutation',
    scale.permutation.importance = TRUE,
    local.importance = TRUE,
    write.forest = TRUE,
    probability = TRUE,
    min.node.size = 2,
    replace = TRUE,
    splitrule = "gini",
    sample.fraction = 0.8154535,
    seed = 12,
    verbose = TRUE,
    oob.error = TRUE,
    classification = TRUE,
    keep.inbag = TRUE
)


print(rf_model_mtbsi_mod)

# Save the model
qs::qsave(rf_model_mtbsi_mod, file = file.path(analyzed_src, filename, "models/rf_mtbsi_models_v2", "mtbsi_mod_pred1.qs"))
qs::qsave(rf_model_mtbsi_mod, file = file.path(analyzed_src, filename, "models/rf_mtbsi_models_v2", "mtbsi_mod_pred4.qs"))

vip::vip(rf_model_mtbsi_mod,
        scale = TRUE, 
        num_features = 20,
        geom = "point", 
        all_permutation = TRUE, 
        aesthetics = list(color = "blue", size = 2.5)) +  # Show top 20 variables
        geom_text(aes(label = sprintf("%.1f", Importance)), 
              hjust = -0.3,  # Adjust horizontal position
              size = 2,    # Text size
              color = "black") +
    # theme_minimal() +
    theme(
        plot.title = element_text(size = 10, face = "bold"),
        axis.text.y = element_text(size = 10),  # Smaller text for many variables
        axis.text.x = element_text(size = 10),
        axis.title = element_text(size = 10),
        plot.margin = margin(1, 1, 1, 1, "cm")
    ) +
    labs(title = "Four Decades Variable Importance Plot for Moderate Fire Severity",
         x = "Variables",
         y = "Variables Importance Score")

# save vip plot
ggsave(file.path(fig_src, "figures_v2/VIP_plot_mtbsi_mod_pred4.png"), 
       dpi = 500, unit = "in", width = 6, height = 8)


rf_model_mtbsi_low <- ranger(
    formula = severity_Low ~ .,
    data = train_data_mtbsi_low,
    num.threads = 4,
    mtry = 2,
    num.trees = 5000,
    importance = 'permutation',
    scale.permutation.importance = TRUE,
    local.importance = TRUE,
    write.forest = TRUE,
    probability = TRUE,
    min.node.size = 2,
    replace = TRUE,
    splitrule = "gini",
    sample.fraction = 0.8154535,
    seed = 12,
    verbose = TRUE,
    oob.error = TRUE,
    classification = TRUE,
    keep.inbag = TRUE
)

print(rf_model_mtbsi_low)

# Save the model
qs::qsave(rf_model_mtbsi_low, file = file.path(analyzed_src, filename, "models/rf_mtbsi_models_v2", "mtbsi_low_pred1.qs"))
qs::qsave(rf_model_mtbsi_low, file = file.path(analyzed_src, filename, "models/rf_mtbsi_models_v2", "mtbsi_low_pred4.qs"))

vip::vip(rf_model_mtbsi_low,
        scale = TRUE, 
        num_features = 20,
        geom = "point", 
        all_permutation = TRUE, 
        aesthetics = list(color = "blue", size = 2.5)) +  # Show top 20 variables
        geom_text(aes(label = sprintf("%.1f", Importance)), 
              hjust = -0.3,  # Adjust horizontal position
              size = 2,    # Text size
              color = "black") +
    # theme_minimal() +
    theme(
        plot.title = element_text(size = 10, face = "bold"),
        axis.text.y = element_text(size = 10),  # Smaller text for many variables
        axis.text.x = element_text(size = 10),
        axis.title = element_text(size = 10),
        plot.margin = margin(1, 1, 1, 1, "pt")
    ) +
    labs(title = "Four Decades Variable Importance Plot for Low Fire Severity",
         x = "Variable",
         y = "Variables Importance Score")

ggsave(file.path(fig_src, "figures_v2/VIP_plot_mtbsi_low_pred4.png"), 
       dpi = 500, unit = "in", width = 6, height = 8)         

```



```{r}
# Random Forest model using randomForest package
train_data_FNF <- train_data_FNF %>%
                    mutate(across(where(is.numeric), ~ round(.x, 2)))


randfor_FNF <- randomForest(event~ ., data = train_data_FNF, ntree = 2000, 
                        mtry = 3, importance = TRUE, localImp = TRUE,
                        keep.inbag = TRUE)


randfor_FNF
# random forest tree
set.seed(432)
randfor_FNF_tree <- randomForest(event~ ., data = train_data_FNF, ntree = 2000, 
                        mtry = 2, importance = TRUE, localImp = TRUE,
                        keep.inbag = TRUE, maxnodes = 20)
print(randfor_FNF_tree)
png(file.path(fig_src,"figure6.png"), width=12, height=8, units="in", res=300)
reprtree:::plot.getTree(randfor_FNF_tree, k=3, depth = 10,
main = "Pruned Random Forest tree of wildfire in western US over four decades", digits = 2)
dev.off()

```


## 9.2 Classification tree
```{r}
library(caret)
library(rpart)      # For decision trees
library(rpart.plot) # For visualization
library(doParallel)

# Setup parallel processing
registerDoParallel(cores = 4)
predictors  <- predictors_collection[[4]]

# Set up train control
ctrl <- trainControl(
  method = "repeatedcv",       
  number = 10,                 
  repeats = 5,                
  allowParallel = TRUE,       # Enable parallel processing
  savePredictions = "final",  # Save predictions
  classProbs = TRUE,          # Return probability predictions
  summaryFunction = twoClassSummary, # For ROC, Sens, Spec
  returnResamp = "all"        # Return all resampling results
)

# Define tuning grid
tuneGrid <- expand.grid(
  cp = seq(0.001, 0.1, by = 0.0001)  
)

# Train the model
tree_cv_fnf <- caret::train(
  formula = as.formula(paste("event", "~", paste(predictors, collapse = " + "))),
  data = train_data_FNF,
  method = "rpart",           
  trControl = ctrl,
  tuneGrid = tuneGrid,
  metric = "ROC"
)

tree_cv_fnf <- train_decision_tree(
    data = train_data_FNF,
    predictors = predictors_collection[[4]],
    response_var = "event"
)






# Train the model
tree_cv_mtbsi <- caret::train(
  severity ~.,
  data = train_data_mtbsi,
  method = "rpart",           
  trControl = ctrl,
  cp = 0.01,
  metric = "ROC"
)

# Print model results
print(tree_cv_fnf)
print(tree_cv_mtbsi)

# Plot the tree
rpart.plot(tree_cv_fnf$finalModel)
rpart.plot(tree_cv_mtbsi$finalModel)

# Variable importance
impfnf <- varImp(tree_cv_fnf)
impmtbsi <- varImp(tree_cv_mtbsi)

plot(impfnf)
plot(impmtbsi)

# Save the plot
# ggsave(file.path(fig_src, "figure6.png"), fimp,
#        width = 7, height = 6, dpi = 300)

# Confusion matrix on test set
predictionsFNF <- predict(tree_cv_fnf, newdata = test_data_FNF)
confusionMatrix(predictionsFNF, test_data_FNF$event)
# Severity
predictionsmtbsi <- predict(tree_cv, newdata = test_data_mtbsi)
confusionMatrix(predictionsmtbsi, test_data_mtbsi$severity)


# ROC curve
library(pROC)
pred_probs <- predict(tree_cv_fnf, newdata = test_data_FNF, type = "prob")
pred_probs
roc_curve <- roc(test_data_FNF$event, pred_probs[,2])
plot(roc_curve)
auc(roc_curve)

# Model evaluation metrics
resamps <- tree_cv_fnf$resample
summary(resamps)

# Visualize cross-validation results
# Plot parameter tuning results
plot(tree_cv_fnf, metric = "ROC", plotType = "level",
     scales = list(x = list(rot = 90)))

# Get best tuning parameters
print(tree_cv_fnf$bestTune); tree_cv_fnf$bestTune$cp

# Save model
saveRDS(tree_cv, "tree_model.rds")

plot(tree_cv)




```

## Pruned classification tree
```{r}
tree_model <- rpart(
  event~.,
  data = train_data_FNF,
  method = "class",
  control = rpart.control(cp = 0.0015) 
)


pruned_tree <- prune(tree_model, cp = 0.004)

# Visualize the pruned tree
rpart.plot(pruned_tree, 
          extra = 101,           
          box.palette = "RdYlGn",
          fallen.leaves = TRUE,   
          type = 3,
          main = "Pruned Classification tree wildfire Occurrence"
          )

ggsave(file.path(fig_src, "figure6.png"),
       width = 7, height = 6, dpi = 500)


```

```{r}

set.seed(368910)
spamBoot1 <- spamApp[
  sample(1:nrow(spamApp), nrow(spamApp), replace = TRUE), ]

libs <- c("rpart", 'randomForest', 'VSURF')
if(!require(libs)) {
  install.packages(libs)
}
lapply(libs, library, character.only = TRUE)

data("Ozone", package = "mlbench")


```